---
title: "Parity and AD specific neurodegeneration"
author: "Clazz"
toc: true
number-sections: true
format: 
  html:
    code-fold: true
editor: source
bibliography: references.bib
csl: elsevier-vancouver.csl
---

```{r, include = F}

if (!require("pacman")) install.packages("pacman")

pacman::p_load(tidyverse, VIM, ggpubr, ggeffects, car, sjPlot,caret, ggrain, emmeans, skimr, lavaan, patchwork, lme4, lmerTest, gtsummary, huxtable, readxl, broom.mixed, pwrss, MuMIn)
```

```{r setup, include=FALSE}

# Project: ASHS_APOE_analysis
# Script: paper_ashs_pregnancy_APOE.R
# Author: Clazz
# Date: 25/09/2025
#
# DATA SOURCES:
# - ALFA+: G:/My Drive/Data/ALFA+
#
#
# Target journal: JPAD
# word limit: 5,500 words, max 5 figures/tables/appendices, max 60 references
# Abtsract: structured abstract (250 words), purpose of research + basic procedures and methods used + main findings with effect sizes and significance+ conclusions and importance
# Highilight:  short collection of bullet points that should capture the novel results of your research as well as any new methods used during your study. Highlights will help increase the discoverability of your article via search engines. 
# Manuscripts must include a statement that all procedures were performed in compliance with relevant laws and institutional guidelines and have been approved by the appropriate institutional committee(s). The statement should contain the date and reference number of the ethical approval(s) obtained.
# Manuscripts must also include a statement that the privacy rights of human subjects have been observed and that informed consent was obtained for experimentation with human subjects.
# Authors must document in the manuscript that ethics committee approval and informed consent have been obtained for studies involving patients or volunteers (including organ/tissue donors). 
#tables can be placed at the end of the article on a separate page
#figures are submitted separately, min 500 dpi, can be saved as PNG
#AMA manual of style
#Article should be divided by numbered sections. subsections may have a brief subheading.


# Register an inline hook:
options(scipen = 1, digits = 3)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 8,
  out.width = "100%",
  dpi = 300
)

config <- config::get()

source("C:/Users/U253662/Documents/git/functions/jt_plot_aesthetics.R")
```

```{r, include=FALSE}
# creating basic dataset

apoe2<-read_delim("G:/My Drive/Data/ALFA+/data_raw/fluid_biomarkers/20201007_APlus_Biomarcadores - corregido.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE)%>%
  dplyr::select(IdParticipante, APOE_consensus)%>%
  rename(subject= IdParticipante)%>%
  mutate(APOE_consensus==as.character(APOE_consensus),
         APOE1 = str_sub(APOE_consensus, start = 1, end = 1),
         APOE2 = str_sub(APOE_consensus, start = 2, end = 2),
         APOE_cat = case_when(APOE1=="4" & APOE2 == "4" ~ "2",
                              APOE1== "4" & APOE2 != "4" ~ "1",
                              APOE1 != "4" & APOE2=="4" ~ "1",
                              .default = "0"))%>%
  dplyr::select(subject, APOE_cat)%>%
  unique()

data_naked<-read_delim(paste0(config$alfa_dataset_cleaned, "ALFA_dataset_basic.csv"), 
    delim = ",", escape_double = FALSE, trim_ws = TRUE)%>%
  mutate(sex=factor(sex),
         APOE_binary = factor(APOE_binary), 
         subject=as.numeric(subject),
         parous_group= factor(parous_group),
         DOB = as.Date (DOB, format = "%Y-%m-%d"),
         kids_dicho=factor(kids_dicho),
         kids_total2 = ifelse(kids_total==4, 3, kids_total))%>%
  dplyr::select(-TIV)%>%
  left_join(., apoe2)%>%
  mutate(kids_total = ifelse(kids_total ==5, 4, kids_total), #group women who have 4 and 5 kids together due to low N
         early_men = ifelse(age_menopause<45, 1, 0))%>%
  filter(age_v1 >= age_menopause | is.na(age_v1) | is.na(age_menopause)) #remove pre-menopause women
  

alfa_subjects<-data_naked%>%
  dplyr::select(subject, sex)

```

```{r hippocampus_data, include = F}
ashs1<-read_excel(paste0(config$alfa_dataset_neuroimaging, "ASHS/bx_ashs_volumes_ALFA_PLUS_V1_20231120_20250702_164509.xlsx"))%>%
  dplyr::select(subject, region, side, volume)%>%
  group_by(subject, region) %>%
  summarize(volume = sum(volume), .groups = "drop")%>%
  pivot_wider(names_from = region, values_from = volume, names_prefix = "vol_")%>%
  mutate(Visit = 1,
         diff_mri = 0,
         subject= as.numeric(subject))

ashs1b<-read_excel(paste0(config$alfa_dataset_neuroimaging, "ASHS/bx_ashs_volumes_ALFA_PLUS_EXP_20250918_161906.xlsx"))%>%
  dplyr::select(subject, region, side, volume)%>%
  group_by(subject, region) %>%
  summarize(volume = sum(volume), .groups = "drop")%>%
  pivot_wider(names_from = region, values_from = volume, names_prefix = "vol_")%>%
  mutate(Visit = 1,
         diff_mri = 0,
         subject= as.numeric(subject))%>%
  filter(subject != 11331,
         subject != 10634,
         subject!= 13368)

date_mri_v1<-read_excel(paste0(config$alfa_dataset_neuroimaging, "scan_dates/bx_scandates_ALFA_PLUS_V1_20231120_20250210_135308.xlsx"))%>%
  dplyr::select(subject_label, scandate)%>%
  mutate(subject_label = as.numeric(subject_label))%>%
  rename(subject = subject_label,
         scandate_v1 = scandate)

date_mri_v2<-read_excel(paste0(config$alfa_dataset_neuroimaging, "scan_dates/bx_scandates_ALFA_PLUS_V2_20230518_20250210_135351.xlsx"))%>%
  dplyr::select(subject_label, scandate)%>%
  mutate(subject_label = as.numeric(subject_label))%>%
  rename(subject = subject_label,
         scandate_v2 = scandate)

diff_mri<-full_join(date_mri_v1, date_mri_v2)%>%
  mutate(scandate_v1 = as.Date(scandate_v1, format = "%Y-%m-%d"),
         scandate_v2 = as.Date(scandate_v2, format = "%Y-%m-%d"),
         diff_mri = as.numeric(difftime(scandate_v2, scandate_v1, units = c("days")))/365.25)%>%
  dplyr::select(subject, diff_mri)
  

ashs2<-read_excel(paste0(config$alfa_dataset_neuroimaging, "ASHS/bx_ashs_volumes_ALFA_PLUS_V2_20230518_20250702_165801.xlsx"))%>%
  dplyr::select(subject, region, side, volume)%>%
  group_by(subject, region) %>%
  summarize(volume = sum(volume), .groups = "drop")%>%
  pivot_wider(names_from = region, values_from = volume, names_prefix = "vol_")%>%
  mutate(Visit = 2,
         subject= as.numeric(subject))%>%
  left_join(., diff_mri)


data_ashs_long<-ashs1%>%
  bind_rows(ashs1b)%>%
  bind_rows(ashs2)%>%
  mutate(subject = as.numeric(subject),
         vol_HIPP = vol_CA1+vol_CA2+vol_CA3+vol_DG+vol_SUB,
         across(starts_with("vol_"), ~./1000))%>%
  right_join(., data_naked)%>%
  filter(subject!= 21136, #this is a very big outlier that increases 1cm3 of HV over visits
         subject!= 12560, #this is a very big outlier that increases 1cm3 of HV over visits
         #subject!= 55854, #cooks distance > 1
         subject != 12186) #is missing CA2 and CA3 values in V1
         #subject != 11154 )#cooks distance > 1
 

# data_ashs_long_outliers<-ashs1%>%
#   bind_rows(ashs1b)%>%
#   bind_rows(ashs2)%>%
#   mutate(subject = as.numeric(subject),
#          vol_HIPP = vol_CA1+vol_CA2+vol_CA3+vol_DG+vol_SUB,
#          across(starts_with("vol_"), ~./1000))%>%
#   right_join(., data_naked)%>%
#   left_join(., A_long)

# data_ashs_long_outliers%>%
#   filter(sex==2)%>%
#   filter(subject== 21136|
#          subject== 12560)%>%
#   mutate(subject=factor(subject))%>%
#   ggplot(aes(x=diff_mri, y=vol_HIPP, colour=subject))+
#   geom_point()+
#   geom_line()+
#   th
# 
# data_ashs_long%>%
#   filter(sex==2)%>%
#   mutate(subject=factor(subject))%>%
#   ggplot(aes(x=Visit, y=vol_HIPP, group=subject))+
#   geom_point()+
#   geom_line()+
#   th
```

```{r, biomarkers, include=F}

biomarkers<-read_delim(paste0(config$alfa_dataset_cleaned, "biomarkers_CSF_processed.csv"), 
    delim = ",", escape_double = FALSE, trim_ws = TRUE)%>%
  mutate(A=factor(A))

biomarkers_time_diff<-read_delim(paste0(config$alfa_dataset_cleaned, "biomarkers_CSF_processed.csv"), 
    delim = ",", escape_double = FALSE, trim_ws = TRUE)%>%
  dplyr::select(subject, collection_dt, Visit)%>%
  pivot_wider(names_from = Visit, values_from = collection_dt)%>%
  mutate(V1 = as.Date(V1, format = "%d/%m/%Y"),
         V2 = as.Date(V2, format = "%d/%m/%Y"),
         biomarkers_time_diff = as.numeric(difftime(V2, V1, units = c("days")))/365.25)%>%
  dplyr::select(subject, biomarkers_time_diff)%>%
  mutate(Visit = 2)
  
data_biomarkers_long<-left_join(data_naked, biomarkers)%>%
  mutate(Visit=ifelse(Visit=="V1", 1, 2))%>%
  left_join(., biomarkers_time_diff)%>%
  mutate(biomarkers_time_diff = ifelse(Visit==1, 0, biomarkers_time_diff))

A_long<-data_biomarkers_long%>%
  dplyr::select(subject, Visit, CSF_Abeta_ratio)
  
```

```{r cognition_data, include=F}
cognition<-read_excel(paste0(config$alfa_dataset_cognition, "PACCv1_to_v3(incomplete).xlsx"))%>%
  dplyr::select(IdParticipante, Visit, Z_PACC, Z_FCSRTI_RT, Z_aWMS_DR_RS, Z_Wclave_totalscr, Z_FS_totalcnswr, FechaResultado)%>%
  rename(subject = IdParticipante)

time_diff<-read_excel(paste0(config$alfa_dataset_cognition, "PACCv1_to_v3(incomplete).xlsx"))%>%
  dplyr::select(IdParticipante, Visit,FechaResultado)%>%
  rename(subject = IdParticipante)%>%
  filter(Visit != 3)%>%
  pivot_wider(names_from = Visit, values_from = FechaResultado, names_prefix = c("Fecha_Resultado"))%>%
  mutate(time_diff = as.numeric(difftime(Fecha_Resultado2, Fecha_Resultado1, units= c("days"))))%>%
  dplyr::select(subject, time_diff)

data_cog_long<- left_join(data_naked, cognition)%>%
  mutate(age_cog = as.numeric(difftime(FechaResultado, DOB, units= c("days")))/365.25)%>%
  group_by(subject)%>%
  filter(Visit != 3) %>%
  filter(n()>=2)%>%
  relocate(Visit, .after=sex)%>%
  ungroup()%>%
  left_join(., time_diff)%>%
  mutate(time_diff = ifelse(Visit == 1, 0, time_diff),
         time_diff = time_diff/365.25)
```

# Abstract

300 words

# Introduction

Alzheimer’s disease (AD) is a neurodegenerative disorder characterised by the accumulation of amyloid-β and tau pathology, leading to neurodegeneration and progressive cognitive and functional decline. Ageing, carrying the Apolipoprotein-E ε4 (APOE ε4) allele, and female sex are three established non-modifiable risk factors for the typical late-onset presentation of this disease @riedel2016. Amongst APOE ε4 carriers, women exhibit an increased risk of progression from healthy ageing to Mild Cognitive Impairment (MCI), or from MCI to AD dementia between the ages of 65 to 75 compared to men [@Altmann2014; @Neu2017]. Mechanisms underlying the interaction between APOE ε4 and sex are not well understood, warranting a deeper investigation of female-sex specific events that might interact with this phenomenon.

In particular, pregnancy is a female-specific event with far reaching biological impacts on multiple bodily functions @oleary1991; @huang2021; @hadden2009, yet its potential effect on AD-related physiological processes remains underexplored. In humans, the first pregnancy is associated with drastic increases in sex steroid levels @brett2001 and a significant hippocampal volume loss @pritschet2024, which only partially recovers in the post-partum period @Hoekzema2017. Whilst no data has been published regarding hippocampal volume change in second time mothers, Straathof and colleagues @straathof2026 recently showed that cortical volume changes that happened during the first pregnancy were mirrored in the second, meaning hippocampal changes might also be reproduced. Interestingly, dynamic grey matter volume changes seen during gestation are associated with the change curve seen in free circulating sex-hormones @servin-barthet2025. In rodents, pregnancy is linked with decreased hippocampal volume during gestation as well as the mid post-partum @duarte-guterman2019. Animal models converge towards the idea that cell proliferation is reduced during pregnancy but returns to nulliparous levels after pups have been weaned @duarte-guterman2019. Together, these findings suggest that pregnancy and the post-partum period affect hippocampal volume dynamically. Therefore, this region seems of particular interest given it is affected by parity and is related to AD neurodegeneration.

The hippocampus is among the first regions affected by AD, and hippocampal atrophy is a hallmark predictor of progression from MCI to AD @apostolova2006. However, few studies have explored the relationship between parity (number of births) and hippocampal volume (HV) in older adult women. One cross-sectional study involving 248 women, ranging from cognitively unimpaired (CU) to those clinically diagnosed with AD, found a significant negative association between grand multiparity (4+ births) and HV @Jung2020. Additionally, preliminary evidence from a study of 1284 CU postmenopausal women from the UK Biobank indicated that women who have experienced at least one birth had lower HV compared to nulliparous women, but this difference was not present when comparing multiparous women to nulliparous women @tucci2025. Interestingly, another study of 19,787 middle-aged CU women from the UK Biobank found that parous women exhibited a significant positive linear association between the hippocampus and the number of childbirths, although this analysis was exploratory @de2020. Additionally, recent evidence from 254 cognitively unimpaired post-menopausal women suggest that the association between hippocampal volume and parity may depend on the presence of early AD neuropathology \[my paper, in press\]. In animal models, nulliparous middle-aged rats show a steeper decline in hippocampal neurogenesis than their primiparous counterparts @eid2019. Together, these findings suggest that past reproductive experience influences hippocampal structure in older age, but this association is inconsistent, meaning other factors may be contributing to and modulating this association.

APOE-e4 may be another modifier of the direction of association between parity and HV in middle age. In a study by Lee and colleagues, wild-type and APOE ε4 models of middle-aged female rats were used to specifically investigate whether APOE status influences the effect of parity on neurogenesis and HV @lee2024. Results showed that primiparous APOE ε4 rats displayed decreased neurogenesis in the ventral dentate gyrus of the hippocampus compared to primiparous wild-type counterparts. In addition to this, parous APOE ε4 rats showed a spatial solving strategy, indicating decreased hippocampal recruitment. Whilst to our knowledge, no study has looked into the potential interplay of APOE-e4 and parity in older women, there is evidence showing that apolipoproteins play a role in pregnancy. A study following 243 women throughout each of their trimester found that maternal concentrations of apolipoproteins change during gestation, and some were linked with metabolic risk factors like liver enzymes, inflammatory markers and blood pressure @chen2025a. In addition to this, in a cohort of 100 pregnant women, APOE-e4 was associated with implantation failure @darabi2024.

This study aims to evaluate whether parity is associated with differential long-term effects on HV and its trajectory as a function of APOE ε4 allele status in postmenopausal CU women at increased risk for AD. To do so, we aimed to test the interactive effects of parity, APOE ε4 carriership and time on hippocampal volume. We hypothesized that APOE ε4 carriers would show a negative relationship between parity and hippocampal volume and trajectory.

# Methods

## Study population

We included postmenopausal women who completed the baseline and first longitudinal visit (with an average time between visits of 3.32 years) from the ongoing ALFA+ study (Clinicaltrials.gov, NCT01835717). ALFA+ is a research cohort of middle-aged cognitively unimpaired participants, many of whom are offspring of AD patients, who have been deeply characterized by clinical interviews, lifestyle and risk factor questionnaires, cognitive testing, CSF biomarkers, and neuroimaging procedures, including magnetic resonance imaging (MRI), and Aβ positron emission tomography (PET). All of these procedures are repeated every 3 years with the main aim of identifying the earliest pathophysiological changes in the preclinical AD continuum @molinuevo2016. Briefly, ALFA+ inclusion criteria were: 1) participants who had previously participated in the 45-65/FPM2012 study (ALFA parent cohort); 2) age between 45 and 75 years at the moment of inclusion in the cohort; 3) long-term commitment to undergo all tests and study procedures (MRI, PET, and lumbar puncture). Exclusion criteria at baseline included: 1) cognitive impairment (Clinical Dementia Rating \[CDR\] \> 0, Mini-Mental State Examination \[MMSE\] \< 26, semantic fluency \< 12); 2) any significant systemic illness or unstable medical condition which could lead to difficulty complying with the protocol; 3) any contraindication to any test or procedure; 4) family history of monogenic AD.

In addition, women with a history of cardiovascular-related gestational disease, such as preeclampsia, were excluded from this analysis. Sex and menopausal status (using STRAW criteria) were self-reported, and women who had not reached menopause were not included.

The study was approved by the Independent Ethics Committee “Parc de Salut Mar”, Barcelona, and all participants gave written informed consent. Baseline visits took place between 2016 and 2019, and the follow-up visit between 2019 and 2022.

## Reproductive variables

The reproductive variables were self-reported by participants during a clinical interview in the baseline visit. Parity was defined as the number of biological children and used as a continuous and dichotomous \[parous vs nulliparous\] variable in the main analyses. Due to the small number of grand multiparous participants, all participants with 4 or more biological children were grouped. Hormonal replacement therapy \[ever used vs not ever used\], cause of menopause \[natural vs non-natural\] and early menopause \[age at menopause\> 45 vs age at menopause≤ 45\] were defined as binary variables. Reproductive span was calculated by subtracting age at menarche from age at menopause. The number of abortions was used as a continuous variable.

## Cognitive measures

The main cognitive outcome of this study was a modified Preclinical Alzheimer’s cognitive composite (mPACC), a sensitive composite measure of early cognitive decline, comprising the total immediate recall of the Free and Cued Selective Reminding Test (FCSRT), the total delayed recall of the logical memory subtest in the Wechsler Memory Scale-IV, the Coding subtest of the Wechsler Adult Intelligence Scale-Fourth Edition (WAIS-IV), and the total score of the semantic fluency test (animals category), as defined in previous work @lópez-martos2025a. Z-scores were computed for Visits 1 and 2 and used in the linear regressions.

## Core AD and glial activation biomarkers

CSF collection, processing, and storage in the ALFA+ study have been described previously @Milà-Alomà2020. Briefly, CSF Aβ42, Aβ40, sTREM2, YKL40, IL-6 and GFAP were measured with the exploratory Roche NeuroToolKit immunoassays (Roche Diagnostics International Ltd, Rotkreuz, Switzerland) on a cobas e 601 module. Measurements were performed at the Clinical Neurochemistry Laboratory, Sahlgrenska University Hospital, Molndal, Sweden. Aß accumulation (Aß+, Aß–) was estimated using CSF Aß42/40. Aß status (Aß+, Aß–) was defined using the cutoff of 0.071 for the ratio Aß42/40 @Milà-Alomà2020 , where any participant below this cutoff was considered to be Aβ+.

## Imaging data acquisition and preprocessing

Additional information on MRI data acquisition is available in the supplementary materials (eMethods 1).

Automatic Segmentation of Hippocampal Subfields (ASHS) software @Yushkevich2015 was used on the T1, T2 and inversion recovery (IR) weighted images to segment the hippocampal formation in the following sub-regions: Brodmann areas 35 and 36, cornu Ammonis (CA) 1, 2 and 3, dentate gyrus (DG), entorhinal cortex (ERC), parahippocampal cortex (PHC), subiculum (SUB), and the hippocampal sulcus (SUL). All segmentations were visually inspected before proceeding with the statistical analyses.

Hippocampal values are reported in cm3 and were obtained at both baseline and follow-up visits.

Total Intracranial volume (TIV) was estimated using deformation fields obtained during the nonlinear registration of each subject’s T1-weighted MRI to a whole-brain template. The FSL Brain Extraction Tool @smith2002 was applied to the template to create an intracranial mask, which was warped into each subject’s MRI at native space to obtain the TIV measurement. TIV is also reported in cm3.

Gray matter volumes were acquired using Freesurfer7's cortical Desikan-Killiany labeling pipeline on the obtained T1 images.

## Statistical analyses

Descriptive statistics were performed using Kruskal-Wallis rank sum test for longitudinal variables and Fisher's exact test for binary variables. Data was grouped by the number of biological children in order to identify any confounders to covary for in the rest of our analyses.

Linear mixed-effects models were used to explore longitudinal associations between exposures and outcomes. This method was chosen over simple linear regression of subject slopes between baseline and follow-up, due to its ability to account for within-subject correlation and accommodate missing data under the missing-at-random (MAR) assumption.

Hippocampal volume was designated as the primary outcome of interest. Cognitive performance was used as a secondary outcome to provide a clinical context to our findings.

Primary analyses assessed the joint effects of parity and APOE-e4 (exposures), on HV (outcome) across study visits, and tested whether this association depended on time. This model included a three-way interaction term among parity, APOE ε4, and time (years between Visit 1 and Visit 2) as well as all lower-order interactions. Analyses were adjusted for age at baseline, education, scaled TIV and HRT. To assess the AD-specificity of our results, the three-way interaction (parity × APOE ε4 × time) was replicated using gray matter volume in the calcarine cortex, a region typically spared in preclinical AD. This model was adjusted for baseline age, education TIV, and HRT.

Sensitivity analyses were performed to determine whether the observed associations were driven by AD-related pathology. Continuous CSF Aβ42/40 load and its interaction with time were introduced as a covariate to disentangle the neuropathological contributions of Aβ accumulation from the APOE ε4-related effects. Given the high correlation between Aβ load and APOE ε4 variables, collinearity was tested for each model (eTable X). In addition to this, we reproduced the analysis in Aβ-negative participants only. Additional sensitivity analyses included additional covariates that reflect estrogen exposure \[early menopause, reproductive span, cause of menopause, number of abortions\] to evaluate their impact on the associations of interest. 

Secondary analyses explored the joint effects of parity and APOE-e4 (exposures) on mPACC (outcome) using the same framework as the primary analysis. Covariates included were for baseline age, HRT and education. Aβ load was then introduced in a second analysis.

Exploratory linear mixed-effects models examined the interaction among parity, APOE-e4, and time on hippocampal subfield volumes across visits. Subfield measures were standardized to enable comparison across regions. Given the exploratory nature of these analyses, no correction for multiple comparisons was applied. Additional exploratory models used CSF inflammatory biomarkers \[GFAP, sTREM2, YKL-40, IL-6\] as outcomes to investigate their potential role in the observed associations. Biomarker values were log-transformed to meet model assumptions and were analyzed both as raw measures and as ratios normalized to Aβ40, since the use of a reference protein may enhance the clinical relevance of CSF biomarkers and be helpful in improving CSF biomarker clinical relevance @karlsson2025.

Statistical significance was set at p \< 0.05. Analyses were conducted using R software, version 4.5.1 (The R Foundation for Statistical Computing, Vienna, Austria). The package ‘emmeans’ was used to calculate estimated marginal mean for direct group comparison of predicted effects @lenth2017. The package tidiverse was used for data processing and plotting @wickham2019. The package gtsummary was used for formatting tables @sjoberg2021. The package ggeffects was used to extract predicted values from regression models @lüdecke2018. The package lme4 was used for performing mixed effect linear regression models @bates2015.

Rates of missing data in exposures or outcome variables at each time point are reported (Table 1).

# Results

## Descriptive results

```{r, descriptives alfa, include = F}
alfa_participants_v1<-read_delim(paste0(config$alfa_dataset_raw, "Alfa Plus Long 2025__AlfaPlus_general_20250304095809(in).csv"), delim= ";", col_names = TRUE)%>%
  rename(subject = IdParticipante)%>%
  right_join(., alfa_subjects)%>%
  filter(sex==2)%>%
  dplyr::select(subject,Age_V1)%>%
  drop_na()%>%
  nrow()

alfa_participants_v2<-read_delim(paste0(config$alfa_dataset_raw, "Alfa Plus Long 2025__AlfaPlus_general_20250304095809(in).csv"), delim= ";", col_names = TRUE)%>%
  rename(subject = IdParticipante)%>%
  right_join(., alfa_subjects)%>%
  filter(sex==2)%>%
  dplyr::select(subject,Age_V2)%>%
  drop_na()%>%
  nrow()

alfa_subjects_dropout<-read_delim(paste0(config$alfa_dataset_raw, "Alfa Plus Long 2025__AlfaPlus_general_20250304095809(in).csv"), delim= ";", col_names = TRUE)%>%
rename(subject = IdParticipante)%>%
  right_join(., alfa_subjects)%>%
  filter(sex==2)%>%
  filter(is.na(Age_V2))%>%
  dplyr::select(subject)

dropout<-read_delim(paste0(config$alfa_dataset_raw, "Alfa Plus Long 2025_2010_STF_20250304095809(in).csv"), delim= ";", skip=1, col_names = TRUE, locale = locale(encoding = "Latin1") )%>%
  rename(subject = IdParticipante)%>%
  right_join(., alfa_subjects_dropout, join_by(subject))%>%
  dplyr::select(subject, STF_state, STF_reas, STF_reas5, STF_EspeIncumCri)%>%
  mutate(STF_reas = factor(STF_reas, levels = c("1", "2", "3", "4", "5", "6", "7"), labels = c("Death", "No response", "Participant withdrawal", "Clinician decision", "Promotor decision", "Protocol issue", "Other")),
         STF_state = factor(STF_state, levels = c("1", "2", "3", "4"), labels = c("Completed", "Failed preescreening", "Failed screening", "Discontinued")))

# cognition1<-read_excel(paste0(config$alfa_dataset_cognition, "Visit 1 - Cognition Z Scores.xlsx"))%>%
#   dplyr::select(-Gender, -Age_V1S1, -Years_Edu, -V1S1_dt, -Z_PACC_Composite_1.2._v1)%>%
#   rename(memory = Z_Memory_Composite_v1, 
#          attention = Z_Attention_Composite_v1,
#          executive = Z_Executive_Composite_v1,
#          language = Z_Language_Composite_v1,
#          visual = Z_Visual_Composite_v1,
#          subject = IdParticipante)%>%
#   mutate(Visit=1)

cognition1<-read_excel("G:/My Drive/Data/ALFA+/data_raw/Cognition/PACCv1_to_v3(incomplete).xlsx")%>%
  rename(subject = IdParticipante)%>%
  group_by(subject)%>%
  filter(n()>1)%>%
  ungroup()%>%
  filter(Visit==1)%>%
  dplyr::select(subject, Z_PACC)
  

ashs_desc<-data_ashs_long%>%
  group_by(subject)%>%
  filter(n()>1)%>%
  ungroup()%>%
  filter(Visit==1)

biomarkers_wide<-biomarkers%>%
  dplyr::select(-A, -AT, -collection_dt, -SampleType)%>%
  pivot_wider(names_from = Visit, values_from = starts_with("CSF_"))

data_table<-data_naked%>%
  filter(sex==2)%>%
  inner_join(., ashs_desc)%>%
  left_join(.,cognition1)%>%
  left_join(.,biomarkers_wide)%>%
  dplyr::select(age_v1, kids_total, APOE_binary, education, vol_tiv, vol_HIPP, Z_PACC, CSF_Abeta_ratio_V1, HRT, early_men, repro_span , cause_menopause,subject)%>%
  mutate(kids_total= ifelse(kids_total==4, "4+", as.character(kids_total)))%>%
  left_join(., diff_mri)%>%
  dplyr::select(age_v1, kids_total, education, APOE_binary, CSF_Abeta_ratio_V1, HRT, early_men, repro_span , cause_menopause, vol_HIPP , diff_mri, Z_PACC, vol_tiv)

desc_table<-tbl_summary(data_table,
                        statistic = list(all_continuous() ~ "{mean} ({sd})"),
            by = kids_total, 
            missing="no",
            label= list(kids_total2 = "Number of childbirths",
                     APOE_binary = "APOE-E4 carriers, Count (%)",
                     age_v1 = "Age at first scan, Mean (SD)",
                     vol_tiv = "Baseline Total Intracranial Volume (cm3), Mean (SD)",
                     diff_mri = "Time between scans, Mean (SD)",
                     education = "Education (years), Mean (SD)",
                     vol_HIPP = "Baseline Hippocampal volume (cm3), Mean (SD)",
                     Z_PACC = "Baseline composite mPACC score, Mean (SD)",
                     CSF_Abeta_ratio_V1 = "CSF Aβ42/40 ratio pg/ml, Mean (SD)",
                     HRT = "HRT users, Count (%)",
                     early_men = "Early menopause, Count (%)",
                     cause_menopause = "Surgical menopause, Count (%)",
                     repro_span = "Reproductive span, Mean (SD)"),
            value = list(APOE_binary= "1",
                         A= "1",
                         HRT = "1",
                         cause_menopause = "non-natural",
                         early_men= "1"))%>%
  add_n()%>%
  add_overall()%>%
  add_p()%>%
  modify_header(label = "**Variables**") %>% # update the column header
  bold_labels()


```

```{r, echo = F, warning=F}
desc_table
```

As shown in Table 1, `r desc_table$table_styling$header$modify_stat_n[9]` (`r desc_table$table_styling$header$modify_stat_n[9]*100/desc_table$table_styling$header$modify_stat_n[8]`%) participants had no previous childbirths, `r desc_table$table_styling$header$modify_stat_n[10]` (`r desc_table$table_styling$header$modify_stat_n[10]*100/desc_table$table_styling$header$modify_stat_n[8]`%) had one previous childbirth, `r desc_table$table_styling$header$modify_stat_n[11]` (`r desc_table$table_styling$header$modify_stat_n[11]*100/desc_table$table_styling$header$modify_stat_n[8]`%) had 2, `r desc_table$table_styling$header$modify_stat_n[12]` (`r desc_table$table_styling$header$modify_stat_n[12]*100/desc_table$table_styling$header$modify_stat_n[8]`%) had 3, and only `r desc_table$table_styling$header$modify_stat_n[13]` (`r desc_table$table_styling$header$modify_stat_n[13]*100/desc_table$table_styling$header$modify_stat_n[8]`%) had 4 or more. Education significantly differed depending on parity.

## Interplay between parity, APOE-E4 and HV decline

```{r, include = F}

data_ashs_female<-data_ashs_long%>%
  filter(sex==2)%>%
  group_by(subject)%>%
  filter(n()==2)%>%
  ungroup()%>%
  left_join(., A_long)%>%
  mutate(A = ifelse(CSF_Abeta_ratio<0.071, 1, 0))

mod_ashs_APOE<-lmer(vol_HIPP ~  kids_total*APOE_binary*diff_mri +age_v1 + scale(vol_tiv) +HRT + education +(1|subject), data = data_ashs_female)
summary(mod_ashs_APOE)
preds <- ggpredict(mod_ashs_APOE, terms = c("diff_mri", "APOE_binary"))

plot(preds)
influenceIndexPlot(mod_ashs_APOE)

#-

###

mod_ashs_APOE_A0<-lmer(vol_HIPP ~  kids_total*diff_mri +age_v1 + scale(vol_tiv) +HRT + education +(1|subject), data = data_ashs_female[data_ashs_female$APOE_binary==0,])
summary(mod_ashs_APOE_A0)

mod_ashs_APOE_A1<-lmer(vol_HIPP ~  kids_total*diff_mri +age_v1 + scale(vol_tiv) +HRT + education +(1|subject), data = data_ashs_female[data_ashs_female$APOE_binary==1,])
summary(mod_ashs_APOE_A1)

mod_ashs_APOE_dicho<-lmer(vol_HIPP ~  kids_dicho*APOE_binary*diff_mri +age_v1 + scale(vol_tiv) + HRT + education +(1|subject), data = data_ashs_female)
summary(mod_ashs_APOE_dicho)
ggpredict(mod_ashs_APOE_dicho, terms = c("APOE_binary", "diff_mri", "kids_dicho")) %>% plot()


##########
  tiv_freesurfer<-read_excel(paste0(config$alfa_dataset_raw, "neuroimaging/cortical thickness/bx_freesurfer7_aparc_ALFA_PLUS_V1_20231120_20250703_110643.xlsx"))%>%
  filter(region=="eTIV")%>%
  dplyr::select(subject, value)%>%
  unique()%>%
  mutate(subject=as.numeric(subject),
         value=as.numeric(value))%>%
  rename(tiv = value)

gray_v1<-read_excel(paste0(config$alfa_dataset_raw, "neuroimaging/cortical thickness/bx_freesurfer7_aparc_ALFA_PLUS_V1_20231120_20250703_110643.xlsx"))%>%
  filter(measurement== "GrayVol")%>%
  dplyr::select(-ID, -measurement)%>%
  pivot_wider(names_from = region, values_from = value, names_prefix = "vol_")%>%
  mutate(across(starts_with("vol_"), ~as.numeric(.)))%>%
  group_by(subject)%>%
  summarise(across(starts_with("vol_"), sum))%>%
  mutate(Visit= "1",
         subject=as.numeric(subject),
         diff_mri= 0)

gray_v2<-read_excel(paste0(config$alfa_dataset_raw, "neuroimaging/cortical thickness/bx_freesurfer7_aparc_ALFA_PLUS_V2_20230518_20250703_111331.xlsx"))%>%
  filter(measurement== "GrayVol")%>%
  dplyr::select(-ID, -measurement)%>%
  pivot_wider(names_from = region, values_from = value, names_prefix = "vol_")%>%
  mutate(across(starts_with("vol_"), ~as.numeric(.)))%>%
  group_by(subject)%>%
  summarise(across(starts_with("vol_"), sum))%>%
  mutate(Visit= "2",
         subject=as.numeric(subject))%>%
  left_join(., diff_mri)

gray_all<-bind_rows(gray_v1, gray_v2)

data_gray<-left_join(data_naked, gray_all)%>%
  filter(sex==2)%>%
  left_join(., tiv_freesurfer)%>%
  group_by(subject)%>%
  filter(n()==2)%>%
  ungroup()%>%
  group_by(subject)%>%
  mutate(gray_all = rowSums(across(starts_with("vol_"))))

mod_cal<-lmer(vol_pericalcarine ~  kids_total*APOE_binary*diff_mri +age_v1 + HRT+scale(tiv) + education  +(1|subject), data = data_gray)
summary(mod_cal)
```

Parity, APOE-ε4, and time showed a significant three-way interaction for total HV (β = `r tidy(mod_ashs_APOE)$estimate[12]`, 95% CI \[`r tidy(mod_ashs_APOE, conf.int = TRUE)$conf.low[12]`, `r tidy(mod_ashs_APOE, conf.int = TRUE)$conf.high[12]`\], p = `r tidy(mod_ashs_APOE)$p.value[12]`), indicating that higher parity was associated with a steeper decline in HV among APOE-ε4 carriers (Figure X). Stratifying by APOE-ε4 status yielded a significant interaction between parity and time in carriers only (β = `r tidy(mod_ashs_APOE_A1)$estimate[8]`, 95% CI \[`r tidy(mod_ashs_APOE_A1, conf.int = TRUE)$conf.low[8]`, `r tidy(mod_ashs_APOE_A1, conf.int = TRUE)$conf.high[8]`\], p \< 0.001) (Figure X). The 3-way interaction between APOE ε4 parity and time was not associated with GMV in the calcarine cortical area (p = `r tidy(mod_cal)$p.value[11]`).

Reproducing the analysis using parity as a dichotomous variable did not yield any significant associaiton with our 3-way estimate. (p = `r tidy(mod_ashs_APOE_dicho)$p.value[12]`).

## Adjusting for potential confounding factors

```{r, include=F}

  

### AB load

mod_ashs_APOE_ab<-lmer(vol_HIPP ~  kids_total*APOE_binary*diff_mri +age_v1 + HRT+ education+scale(vol_tiv) + CSF_Abeta_ratio + CSF_Abeta_ratio:diff_mri  +(1|subject), data = data_ashs_female)
summary(mod_ashs_APOE_ab)

mod_ashs_APOE_ab_a0<-lmer(vol_HIPP ~  kids_total*diff_mri +age_v1 + scale(vol_tiv) + CSF_Abeta_ratio+ CSF_Abeta_ratio:diff_mri+ HRT+ education  +(1|subject), data = data_ashs_female[data_ashs_female$APOE_binary==0,])
summary(mod_ashs_APOE_ab_a0)

mod_ashs_APOE_ab_a1<-lmer(vol_HIPP ~  kids_total*diff_mri +age_v1 + scale(vol_tiv) + CSF_Abeta_ratio+ CSF_Abeta_ratio:diff_mri+ education  +(1|subject), data = data_ashs_female[data_ashs_female$APOE_binary==1,])
summary(mod_ashs_APOE_ab_a1)

mod_ashs_APOE_abneg<-lmer(vol_HIPP ~  kids_total*APOE_binary*diff_mri +age_v1 + HRT+ CSF_Abeta_ratio+ CSF_Abeta_ratio:diff_mri+ education+scale(vol_tiv)   +(1|subject), data = data_ashs_female[data_ashs_female$A==0,])
summary(mod_ashs_APOE_abneg)



anova_data<-data_ashs_female%>%
  drop_na(repro_span, early_men, cause_menopause)

mod_ashs_APOE_horm<-lmer(vol_HIPP ~  kids_total*APOE_binary*diff_mri +age_v1 + scale(vol_tiv)  + education+ repro_span+ early_men+ HRT+ cause_menopause + (1|subject), data = anova_data)
summary(mod_ashs_APOE_horm)

anova_mod<-lmer(vol_HIPP ~  kids_total*APOE_binary*diff_mri +age_v1 + scale(vol_tiv) +HRT + education+(1|subject), data = anova_data)

anova(mod_ashs_APOE_horm, anova_mod)
```

To account for potential confounding factors, the primary analysis was repeated including Aβ load as a covariate. The triple interaction remained significant after adjustment for Aβ (β = `r tidy(mod_ashs_APOE_ab)$estimate[14]`, 95% CI \[`r tidy(mod_ashs_APOE_ab, conf.int = TRUE)$conf.low[14]`, `r tidy(mod_ashs_APOE_ab, conf.int = TRUE)$conf.high[14]`\], p = `r tidy(mod_ashs_APOE_ab)$p.value[14]`). Aβ load itself showed a strong association with the longitudinal outcome (β = `r tidy(mod_ashs_APOE_ab)$estimate[13]`, 95% CI \[`r tidy(mod_ashs_APOE_ab, conf.int = TRUE)$conf.low[13]`, `r tidy(mod_ashs_APOE_ab, conf.int = TRUE)$conf.high[13]`\], p = `r tidy(mod_ashs_APOE_ab)$p.value[13]`), suggesting a potential independent contribution.

When reproducing the analysis including only participants with negative Aβ status, the effect was maintained, although attenuated (CSF Aβ\<0.071 pg/ml, n = 128) (β = `r tidy(mod_ashs_APOE_abneg)$estimate[14]`, 95% CI \[`r tidy(mod_ashs_APOE_abneg, conf.int = TRUE)$conf.low[14]`, `r tidy(mod_ashs_APOE_abneg, conf.int = TRUE)$conf.high[14]`\], p = `r tidy(mod_ashs_APOE_abneg)$p.value[14]`).

The analysis was further adjusted for hormonal variables (HRT use, reproductive span, early menopause, and cause of menopause), and the three-way interaction remained significant (β = `r tidy(mod_ashs_APOE_horm)$estimate[15]`, 95% CI \[`r tidy(mod_ashs_APOE_horm, conf.int = TRUE)$conf.low[15]`, `r tidy(mod_ashs_APOE_horm, conf.int = TRUE)$conf.high[15]`\], p = `r tidy(mod_ashs_APOE_horm)$p.value[15]`). Additionally, model comparison using an ANOVA test indicated that inclusion of these proxy variables did not improve model fit relative to the original model.

## Interplay between parity, APOE-e4 and cognitive trajectory

```{r, include=F}

cognition<-read_excel("G:/My Drive/Data/ALFA+/data_raw/Cognition/PACCv1_to_v3(incomplete).xlsx")%>%
  dplyr::select(IdParticipante, Visit, Z_PACC, Z_FCSRTI_RT, Z_aWMS_DR_RS, Z_Wclave_totalscr, Z_FS_totalcnswr, FechaResultado)%>%
  rename(subject = IdParticipante)

time_diff<-read_excel("G:/My Drive/Data/ALFA+/data_raw/Cognition/PACCv1_to_v3(incomplete).xlsx")%>%
  dplyr::select(IdParticipante, Visit,FechaResultado)%>%
  rename(subject = IdParticipante)%>%
  filter(Visit != 3)%>%
  pivot_wider(names_from = Visit, values_from = FechaResultado, names_prefix = c("Fecha_Resultado"))%>%
  mutate(time_diff = as.numeric(difftime(Fecha_Resultado2, Fecha_Resultado1, units= c("days"))))%>%
  dplyr::select(subject, time_diff)

data_cog_long<- data_naked %>%
  left_join(.,cognition)%>%
  filter(sex==2)%>%
  mutate(age_cog = as.numeric(difftime(FechaResultado, DOB, units= c("days")))/365.25)%>%
  group_by(subject)%>%
  filter(Visit != 3) %>%
  relocate(Visit, .after=sex)%>%
  ungroup()%>%
  left_join(., time_diff)%>%
  mutate(time_diff = ifelse(Visit == 1, 0, time_diff),
         time_diff_years = time_diff/365.25)%>%
  mutate(Visit=as.numeric(Visit))%>%
  left_join(., A_long)%>%
  group_by(subject)%>%
  filter(n()>1)%>%
  ungroup()



fit_cog<-lmer(Z_PACC ~ time_diff_years*kids_total*APOE_binary +age_v1 + HRT +education +(1 | subject), data_cog_long)
summary(fit_cog)
fit_cog_ab<-lmer(Z_PACC ~ time_diff_years*kids_total*APOE_binary +age_v1 + HRT +education + CSF_Abeta_ratio+ CSF_Abeta_ratio:time_diff_years+ (1 | subject), data_cog_long)
summary(fit_cog_ab)
influenceIndexPlot(fit_cog_ab)
plot(fit_cog_ab)
plot_model(fit_cog_ab, type="int")

#chose not to remove because appart from cook's distance everything seems fine.. fine in other domains too. But she is driving the results a lot so could report with and without?

ggplot(data_cog_long, aes(x= time_diff, y= Z_PACC, group = subject, colour = APOE_binary))+
  geom_line()+
  facet_wrap(~kids_total)
```

The interaction between parity, APOE ε4 and time did not yield a significant association with PACCm (p = `r tidy(fit_cog)$p.value[11]`). Adjusting the analysis for AB strengthened the lower-order interaction between APOE ε4 and parity but it still did not reach signifiance (β= `r tidy(fit_cog_ab)$estimate[11]`, 95% CI \[`r tidy(fit_cog_ab, conf.int = TRUE)$conf.low[11]`, `r tidy(fit_cog_ab, conf.int = TRUE)$conf.high[10]`\], p= `r tidy(fit_cog_ab)$p.value[11]`).

## Exploratory analysis of the interplay between parity, APOE-e4 and hippocampal subfields

```{r, include = F, warning=FALSE}

CA1_mean<-mean(data_ashs_female$vol_CA1[data_ashs_female$Visit==1], na.rm=T)
CA1_sd<-sd(data_ashs_female$vol_CA1[data_ashs_female$Visit==1], na.rm=T)
CA2_mean<-mean(data_ashs_female$vol_CA2[data_ashs_female$Visit==1], na.rm=T)
CA2_sd<-sd(data_ashs_female$vol_CA2[data_ashs_female$Visit==1], na.rm=T)
CA3_mean<-mean(data_ashs_female$vol_CA3[data_ashs_female$Visit==1], na.rm=T)
CA3_sd<-sd(data_ashs_female$vol_CA3[data_ashs_female$Visit==1], na.rm=T)
DG_mean<-mean(data_ashs_female$vol_DG[data_ashs_female$Visit==1], na.rm=T)
DG_sd<-sd(data_ashs_female$vol_DG[data_ashs_female$Visit==1], na.rm=T)
SUB_mean<-mean(data_ashs_female$vol_SUB[data_ashs_female$Visit==1], na.rm=T)
SUB_sd<-sd(data_ashs_female$vol_SUB[data_ashs_female$Visit==1], na.rm=T)

data_ashs_female<-data_ashs_female%>%
  mutate(vol_CA1_std = (vol_CA1 - CA1_mean) / CA1_sd,
         vol_CA2_std = (vol_CA2 - CA2_mean) / CA2_sd,
         vol_CA3_std = (vol_CA3 - CA3_mean) / CA3_sd,
         vol_DG_std = (vol_DG - DG_mean) / DG_sd,
         vol_SUB_std = (vol_SUB - SUB_mean) / SUB_sd)

mod_ashs_CA1<-lmer(vol_CA1_std ~  kids_total*diff_mri*APOE_binary +age_v1  + HRT+ education+scale(vol_tiv)  +(1|subject), data = data_ashs_female)
summary(mod_ashs_CA1)
influenceIndexPlot(mod_ashs_CA1)
mod_ashs_CA1_a1<-lmer(vol_CA1_std ~  kids_total*diff_mri +age_v1  + HRT+ education+scale(vol_tiv)  +(1|subject), data = data_ashs_female[data_ashs_female$APOE_binary==1,])
summary(mod_ashs_CA1_a1)
mod_ashs_CA1_ab<-lmer(vol_CA1_std ~  kids_total*diff_mri*APOE_binary +age_v1  + CSF_Abeta_ratio+ CSF_Abeta_ratio:diff_mri+ HRT+ education+scale(vol_tiv)  +(1|subject), data = data_ashs_female)
summary(mod_ashs_CA1_ab)

mod_ashs_CA2<-lmer(vol_CA2_std ~ kids_total*diff_mri*APOE_binary + age_v1  + HRT+ education+scale(vol_tiv)  +(1|subject), data = data_ashs_female)
summary(mod_ashs_CA2)
influenceIndexPlot(mod_ashs_CA2)
mod_ashs_CA2_a1<-lmer(vol_CA2_std ~ kids_total*diff_mri + age_v1  + HRT+ education+scale(vol_tiv)  +(1|subject), data = data_ashs_female[data_ashs_female$APOE_binary==1,])
summary(mod_ashs_CA2_a1)
mod_ashs_CA2_ab<-lmer(vol_CA2_std ~ kids_total*diff_mri*APOE_binary + age_v1  + HRT+ education+scale(vol_tiv) + CSF_Abeta_ratio + CSF_Abeta_ratio:diff_mri+(1|subject), data = data_ashs_female)
summary(mod_ashs_CA2_ab)

mod_ashs_CA3<-lmer(vol_CA3_std ~  kids_total*diff_mri*APOE_binary +age_v1  + HRT+ education+scale(vol_tiv)  +(1|subject), data = data_ashs_female)
summary(mod_ashs_CA3)
influenceIndexPlot(mod_ashs_CA3)

mod_ashs_DG<-lmer(vol_DG_std ~  kids_total*diff_mri*APOE_binary +age_v1  + HRT+ education+scale(vol_tiv)  + (1|subject), data = data_ashs_female)
summary(mod_ashs_DG)
influenceIndexPlot(mod_ashs_DG)
mod_ashs_DG_a1<-lmer(vol_DG_std ~  kids_total*diff_mri +age_v1  + HRT+ education+scale(vol_tiv)  + (1|subject), data = data_ashs_female[data_ashs_female$APOE_binary==1,])
summary(mod_ashs_DG_a1)
mod_ashs_DG_ab<-lmer(vol_DG_std ~  kids_total*diff_mri*APOE_binary +age_v1  + HRT+ education+scale(vol_tiv)  + CSF_Abeta_ratio + CSF_Abeta_ratio:diff_mri+ (1|subject), data = data_ashs_female)
summary(mod_ashs_DG_ab)

mod_ashs_SUB<-lmer(vol_SUB_std ~  kids_total*diff_mri*APOE_binary +age_v1  + HRT+ education+scale(vol_tiv)  +(1|subject), data = data_ashs_female)
summary(mod_ashs_SUB)
influenceIndexPlot(mod_ashs_SUB)


```

The 3-way interaction between parity, APOE-ε4 and time was significant in the CA1 (β= `r tidy(mod_ashs_CA1)$estimate[12]`, 95% CI \[`r tidy(mod_ashs_CA1, conf.int = TRUE)$conf.low[12]`, `r tidy(mod_ashs_CA1, conf.int = TRUE)$conf.high[12]`\], p= `r tidy(mod_ashs_CA1)$p.value[12]`), CA2 (β= `r tidy(mod_ashs_CA2)$estimate[12]`, 95% CI \[`r tidy(mod_ashs_CA2, conf.int = TRUE)$conf.low[12]`, `r tidy(mod_ashs_CA2, conf.int = TRUE)$conf.high[12]`\], p= `r tidy(mod_ashs_CA2)$p.value[12]`) and in the DG (β= `r tidy(mod_ashs_DG)$estimate[12]`, 95% CI \[`r tidy(mod_ashs_DG, conf.int = TRUE)$conf.low[12]`, `r tidy(mod_ashs_DG, conf.int = TRUE)$conf.high[12]`\], p= `r tidy(mod_ashs_DG)$p.value[12]`) where similarly to the analysis in total HV, parity was associated with steeper HV decline in APOE-ε4 carriers and the opposite in non-carriers. When stratifying the analysis by APOE-ε4 status, only the carrier group remained significant in the CA1 (β= `r tidy(mod_ashs_CA1_a1)$estimate[8]`, 95% CI \[`r tidy(mod_ashs_CA1_a1, conf.int = TRUE)$conf.low[8]`, `r tidy(mod_ashs_CA1_a1, conf.int = TRUE)$conf.high[8]`\], p= `r tidy(mod_ashs_CA1_a1)$p.value[8]`), CA2 (β= `r tidy(mod_ashs_CA2_a1)$estimate[8]`, 95% CI \[`r tidy(mod_ashs_CA2_a1, conf.int = TRUE)$conf.low[8]`, `r tidy(mod_ashs_CA2_a1, conf.int = TRUE)$conf.high[8]`\], p= `r tidy(mod_ashs_CA2_a1)$p.value[8]`) and DG (β= `r tidy(mod_ashs_DG_a1)$estimate[8]`, 95% CI \[`r tidy(mod_ashs_DG_a1, conf.int = TRUE)$conf.low[8]`, `r tidy(mod_ashs_DG_a1, conf.int = TRUE)$conf.high[8]`\], p= `r tidy(mod_ashs_DG_a1)$p.value[8]`).

Adjusting whole group analysis by AB increased effects in the CA1 (β= `r tidy(mod_ashs_CA1_ab)$estimate[14]`, 95% CI \[`r tidy(mod_ashs_CA1_ab, conf.int = TRUE)$conf.low[14]`, `r tidy(mod_ashs_CA1_ab, conf.int = TRUE)$conf.high[14]`\], p= `r tidy(mod_ashs_CA1_ab)$p.value[14]`) but decreased them in the CA2 (p= `r tidy(mod_ashs_CA2_ab)$p.value[14]`). Effects in the DG remained largely unchanged (β= `r tidy(mod_ashs_DG_ab)$estimate[14]`, 95% CI \[`r tidy(mod_ashs_DG_ab, conf.int = TRUE)$conf.low[14]`, `r tidy(mod_ashs_DG_ab, conf.int = TRUE)$conf.high[14]`\], p= `r tidy(mod_ashs_DG_ab)$p.value[14]`).

## Exploratory analysis for inflammatory biomarkers

```{r, include=F}
data_biomarkers_long_female<-data_biomarkers_long%>%
  filter(sex==2)%>%
  group_by(subject)%>%
  filter(n()>1)%>%
  ungroup()%>%
  left_join(., A_long)

data_biomarkers_long_female_v1<-data_biomarkers_long_female%>%
  filter(Visit==1, 
         subject!= "44214")%>%
  mutate(strem_ratio= CSF_sTREM2/CSF_Abeta40,
         ykl_ratio= CSF_YKL40/CSF_Abeta40,
         gfap_ratio = CSF_GFAP/CSF_Abeta40)

fit_base_strem<- lm(log(CSF_GFAP) ~ kids_total*APOE_binary + age_v1 + education, data= data_biomarkers_long_female_v1)
summary(fit_base_strem)
plot(fit_base_strem)
```

```{r, eval=F, include=F}
library(FactoMineR)
library(factoextra)



data_PCA_inf<-data_biomarkers_long_female%>%
  dplyr::select(subject, Visit, CSF_GFAP, CSF_sTREM2, CSF_YKL40, biomarkers_time_diff)%>%
  pivot_wider(names_from = Visit, values_from = c(CSF_GFAP, CSF_sTREM2, CSF_YKL40, biomarkers_time_diff))%>%
  mutate(diff_CSF_GFAP = (CSF_GFAP_2 - CSF_GFAP_1)/biomarkers_time_diff_2,
         diff_CSF_sTREM2 = (CSF_sTREM2_2 - CSF_sTREM2_1)/biomarkers_time_diff_2,
         diff_CSF_YKL40 = (CSF_YKL40_2 - CSF_YKL40_1)/biomarkers_time_diff_2)%>%
  dplyr::select(diff_CSF_GFAP, diff_CSF_sTREM2, diff_CSF_YKL40)


pca_inf<-PCA(data_PCA_inf)
eigenvalues_inf <- pca_inf$eig
eigenvalues_inf

barplot(eigenvalues_inf[, 2], names.arg=1:nrow(eigenvalues_inf), 
       main = "Variances",
       xlab = "Principal Components",
       ylab = "Percentage of variances",
       col ="steelblue")
# Add connected line segments to the plot
lines(x = 1:nrow(eigenvalues_inf), eigenvalues_inf[, 2], 
      type="b", pch=19, col = "red")


fviz_pca_var(pca_inf, col.var="contrib")

pc1<-pca_inf$ind$coord
data_biomarkers_long_female_pc1<-data_biomarkers_long_female%>%
  group_by(subject)%>%
  filter(n()>1)%>%
  slice_head()%>%
  ungroup()

data_biomarkers_long_female_pc1$pc1<-pc1[,1]

fit_pca_inf<-lm(pc1 ~ kids_total*APOE_binary+ age_v1, data = data_biomarkers_long_female_pc1)
plot(fit_pca_inf)
assumptions(fit_pca_inf)


summary(trafo_lm(fit_pca_inf, trafo = "gpower"))

pca_inf$var$coord[,1]

```

No significant effect of the interaction between parity and APOE-e4 was found with any of the tested outcomes.

# Discussion (work in progress)

Our study aimed to investigate the impact of parity on hippocampal volume in CU post-menopausal women at risk of AD, and how APOE-ε4 modified this association. Parity and APOE-ε4 interacted to impact hippocampal volume change in our sample significantly. Stratified analyses revealed that this effect was driven by APOE-ε4 carriers, who showed steeper HV loss with higher parity, suggesting that genetic factors may modify the effect of reproductive history on HV. These interactive effects remained significant after adjustment for potential confounding factors, such as Aβ accumulation and other reproductive variables. Together, these findings highlight that parity may exert a long-term effect on HV in postmenopausal women, but this is dependent on genetic predisposition.

Interestingly, results were not maintained when parity was treated as a dichotomous variable (nulliparous vs parous), suggesting the observed phenomenon may be dose-dependent. Although APOE-ε4 carriers showed significantly higher accumulations of Aβ load, adjusting for this variable did not seem to explain the observed effect. Moreover, this effect was also observed when the same analysis was repeated with Aβ-negative individuals only. Recent evidence from the same cohort highlighted a possible modifying effect of amyloid status on the association between parity and HV, but this was not time-dependent. Therefore, our results do not seem to reflect amyloid-driven neurodegeneration, but instead an independent pathway specific to APOE-ε4 and parity.

We did not find a significant three-way interaction among parity, APOE-ε4, and time in predicting cognitive change. This dissociation between our structural and cognitive outcomes may be due to the relatively young age of our participants or the relatively short follow-up interval. Notably, hippocampal volume loss can precede detectable cognitive decline by years in preclinical AD populations @jack2013, and it is possible that our sample had not yet reached the threshold at which structural alterations manifest as measurable cognitive impairment. However, these changes may also reflect a change in HV separate from neurodegeneration, potentially arising from interindividual variability or nonpathological aging-related factors. Adjusting for amyloid load did strengthen the lower-order association between parity and APOE-ε4; consistent with the HV findings, carriers showed worse cognition with higher parity. However, this did not reach significance. Interestingly, Lee and colleagues observed a shift toward non-spatial cognitive strategies in primiparous APOE-ε4 dams, suggesting reduced hippocampal use @lee2024. These results lend some support to the idea that such effects may eventually emerge in our cohort. Interestingly, our previous work found that parity and amyloid status together predicted dementia-related cognitive change, further suggesting that the pathway linking parity and APOE-ε4 to hippocampal volume loss observed here may operate independently of amyloid-driven decline [in press]. Additional time points will be essential to determine the clinical significance of the structural changes.

Exploratory analyses of hippocampal subfields suggested that CA1 and DG drive the effect observed at the whole-hippocampal level. Similar to whole hippocampal analyses, stratified analyses in APOE-ε4 subgroups showed that the effect was driven by carriers of the allele. Our findings mirror results from Lee and colleagues, who found that primiparous wildtype rats showed a higher number of immature neurons in the ventral DG compared to nulliparous, but that nulliparous APOE-ε4 rats showed a higher number of immature neurons in that area than their primiparous counterparts @lee2024. Notably, the CA1 is an impacted area of the hippocampus during pregnancy in humans @pritschet2024 and rodents show changes both in the CA1 @puri2023 and DG @pawluski2007a. However, these changes are highly dynamic and unlikely to reflect neurodegeneration or persist into older age @servin-barthet2025a. One possible explanation for these results could be that parity may interact with prior genetic vulnerability to affect “inflammageing” @zavatta2022. It is worth noting that sensitivity analyses adjusting for amyloid load did not impact the association between our predictor and CA1 volume, even though this subfield is generally involved in AD-related neurodegeneration @deflores2015. In brief, these findings do not seem to reflect amyloid-dependent atrophy but may point toward an interaction of pregnancy and genetic vulnerability that could contribute to an accelerated senescence. Whether this interaction extends to further genetic factors beyond APOE will be investigated in our cohort.

The exposures observed here operate on different timescales. Pregnancy is a discrete event, concentrated within a specific and distant time window in our cohort, while APOE-ε4 carriership represents a lifelong genetic vulnerability whose effects may compound with age. This asymmetry makes it difficult to determine whether the hippocampal changes observed here reflect acute neurobiological shifts occurring at the time of the reproductive event, or rather the cumulative consequence of long-term physiological changes set in motion by pregnancy and unfolding gradually over decades. One possible explanation is Brinton’s “healthy cell bias” theory may help explain these findings by proposing that estrogen is neuroprotective only under conditions of preserved neuronal metabolic health @brinton2008. APOE-ε4 carriers may represent a metabolically vulnerable group, potentially increasing susceptibility to the hormonal and oxidative changes associated with pregnancy @chen2025b. Although originally developed in the context of hormone therapy, this framework may also apply to parity-related brain change @yu2025a. However, further studies examining the long-term effects of pregnancy on human subjects are needed to fully confirm this theory.


Our data did not find a significant association between inflammation-related CSF biomarkers and the interaction among APOE-ε4, parity, and time. Inflammatory processes serve essential physiological roles in parturition @Challis2009, yet their long-term consequences remain understudied in humans. Neuroinflammation is a known contributor of neurodegeneration @ransohoff2016,  and the APOE-ε4 genotype is linked with a chronic pro-inflammatory state @tao2018. Inflammatory pathways offered a plausible mechanistic explanation for the main findings. Interestingly, Lee and colleagues found that primiparous middle-aged hAPOE4 rats showed a greater inflammatory phenotype compared to wild-type rats in the ventral hippocampus @lee2024, suggesting that APOE4 might also modify parity-related long-term inflammatory responses. The absence of significant findings in our sample might reflect the limited precision of CSF measurements relative to region-specific measurements that are possible on animal models. Larger samples and additional time points will be needed to better characterise longitudinal inflammatory patterns in older parous and nulliparous women.

Some limitations should be considered. Firstly, the number of follow-up visits is insufficient to observe nonlinear evolution, therefore limiting our understanding of possible mechanisms underlying AD pathogenesis. Although the present findings provide an important characterisation of hippocampal trajectories at the preclinical stage, where early detection has the greatest potential clinical relevance, additional follow-up visits are underway in our cohort to better characterise trajectory. Secondly,  our cohort is not diverse: the vast majority of ALFA+ study participants are white Caucasians, primarily from middle- to upper-middle-class backgrounds, resulting in an ethnically homogeneous sample. Future work should reproduce these analyses in a diverse cohort to understand the impact of ethnicity and social background on the present results. In addition to this, our measures are far removed from pregnancy and parturition in time, which leaves at least a couple of decades of brain changes unobserved. Additional work is needed to better bridge the gap between pregnancy and the menopause transition, to understand the impact of parity on grey matter volume trajectories over the lifespan. Lastly, the absence of direct hormonal measurements limits our ability to draw definitive conclusions about the role of sex steroid hormones in the observed associations. By relying on proxy measures, we can only estimate hormone-related effects, potentially leading to an underestimation of their actual influence. 

To conclude, this research showed preliminary evidence that parity is associated with HV trajectory in postmenopausal women depending on their genetic profile. Specifically, APOE-ε4 carriers showed steeper hippocampal volume loss with higher parity, an effect that was dose-dependent, subfield-specific, and independent of amyloid accumulation. These findings suggest that the long-term neurobiological consequences of pregnancy are not uniform across individuals, but are instead shaped by underlying genetic vulnerability. Taken together with the broader literature on sex-specific risk factors for AD, our results underscore the importance of incorporating reproductive history into models of brain aging in women. Future work with longer follow-up, larger and more diverse samples, and direct hormonal measurements will be essential to clarify the mechanisms at play and determine whether the structural changes observed here carry clinical significance over time.

# Figures

## Figure 1

```{r, echo=F, warning=F}
# Create grid of values for prediction - specify multiple time points

individual_slopes <- data_ashs_female %>%
  filter(!is.na(kids_total) & !is.na(APOE_binary) & !is.na(diff_mri) & !is.na(vol_HIPP)) %>%
  group_by(subject, kids_total, APOE_binary) %>%  # Only keep subjects with at least 2 observations
  summarise(
    individual_slope = coef(lm(vol_HIPP ~ diff_mri))[2],
    n_obs = n(),
    .groups = 'drop'
  ) %>%
  filter(!is.na(individual_slope))%>%
  mutate(APOE_binary = as.factor(APOE_binary))



time_range <- range(data_ashs_long$diff_mri, na.rm = TRUE)
time_seq <- seq(from = time_range[1], to = time_range[2])

emm_grid <- emmeans(mod_ashs_APOE, ~ diff_mri * kids_total * APOE_binary,
                    at = list(time = time_seq),
                    data= data_ashs_long)
emm_df <- as.data.frame(emm_grid)

slopes <- emtrends(mod_ashs_APOE, ~ kids_total * APOE_binary, var = "diff_mri", data= data_ashs_long, at = list(kids_total = c(0, 1, 2, 3, 4)))
slopes_df <- as.data.frame(slopes)

p1<-ggplot(slopes_df, aes(x = kids_total, y = diff_mri.trend, 
                            fill = as.factor(APOE_binary))) +
  geom_col(position = "dodge", alpha = 0.8, color = "black", size = 0.3) +
  geom_errorbar(aes(ymin = diff_mri.trend - SE, ymax = diff_mri.trend + SE),
                position = position_dodge(width = 0.9), width = 0.3)+
  th+
  labs(x = "Number of biological children", 
     y = "Time slope", 
     color = expression(italic("APOE")*"-ε4"),
     fill = expression(italic("APOE")*"-ε4"))+
  geom_text(data = data.frame(parous_group = factor("Multiparous", 
                                                     levels = c("Nulliparous", "Primiparous", "Multiparous")), 
                              time_diff = Inf, 
                              Z_PACC = Inf, 
                              label = "p= 0.003**"),
            aes(x = time_diff, y = Z_PACC, label = label),
            size = 4, vjust = 1, hjust = 1, inherit.aes = FALSE)+
 geom_point(data = individual_slopes, 
             aes(x = kids_total, y = individual_slope, 
                 color = as.factor(APOE_binary)),
             position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.9),
             alpha = 0.4, size = 2, shape=21, colour="black") +
  scale_fill_manual(values = c("#1CA8B7", "#FF7043"),
                     labels = c("Non-Carriers", "Carriers"))+
scale_color_manual(values = c("#1CA8B7", "#FF7043"),
                     labels = c("Non-Carriers", "Carriers"))+
  scale_x_continuous(
    breaks = c(0, 1, 2, 3, 4),
    labels = c("0", "1", "2", "3", "4+")
  )

```

```{r, echo=F, warning=F}

#install.packages("ggtext")
library(ggtext)


# Get slopes across parity values for each APOE group
parity_range <- c(0, 1, 2, 3, 4)

# Non-carriers
slopes_nc <- emtrends(mod_ashs_APOE_A0,
                      specs = ~ kids_total,
                      var = "diff_mri",
                      at = list(kids_total = parity_range),
                      data=data_ashs_female[data_ashs_female$APOE_binary==0,])
slopes_nc_df <- as.data.frame(slopes_nc)
slopes_nc_df$APOE_binary <- "Non-carriers"
slopes_nc_df$cov<-0

# Carriers  
slopes_c <- emtrends(mod_ashs_APOE_A1,
                     specs = ~ kids_total,
                     var = "diff_mri",
                     at = list(kids_total = parity_range),
                     data=data_ashs_female[data_ashs_female$APOE_binary==1,])
slopes_c_df <- as.data.frame(slopes_c)
slopes_c_df$APOE_binary <- "Carriers"
slopes_c_df$cov<-0

# Non-carriers + amyloid
slopes_nc_ab <- emtrends(mod_ashs_APOE_ab_a0,
                      specs = ~ kids_total,
                      var = "diff_mri",
                      at = list(kids_total = parity_range),
                      data=data_ashs_female[data_ashs_female$APOE_binary==0,])
slopes_nc_ab_df <- as.data.frame(slopes_nc_ab)
slopes_nc_ab_df$APOE_binary <- "Non-carriers"
slopes_nc_ab_df$cov<-1

# Carriers  + amyloid
slopes_c_ab <- emtrends(mod_ashs_APOE_ab_a1,
                     specs = ~ kids_total,
                     var = "diff_mri",
                     at = list(kids_total = parity_range),
                     data=data_ashs_female[data_ashs_female$APOE_binary==1,])
slopes_c_ab_df <- as.data.frame(slopes_c_ab)
slopes_c_ab_df$APOE_binary <- "Carriers"
slopes_c_ab_df$cov<-1

# Combine
all_slopes <- rbind(slopes_nc_df, slopes_c_df)
all_slopes_cov<- rbind(slopes_nc_ab_df, slopes_c_ab_df)

# Same data as above, but on one plot
p2<-all_slopes%>%
  mutate(APOE_binary=factor(APOE_binary, levels = c("Non-carriers", "Carriers")))%>%
ggplot( aes(x = kids_total, y = diff_mri.trend)) +
  geom_ribbon(aes(ymin = lower.CL, ymax = upper.CL, fill = APOE_binary), 
              alpha = 0.3) +
  geom_ribbon(data = all_slopes_cov, aes(ymin = lower.CL, ymax = upper.CL, fill = APOE_binary), 
              alpha = 0.3)+
  geom_line(aes(color = APOE_binary, linetype = "Unadjusted"), size = 1) +
  geom_line(data = all_slopes_cov, aes(x = kids_total, y = diff_mri.trend, color = APOE_binary,  linetype = "Adjusted"), size = 1, inherit.aes = F)+
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  facet_wrap(~APOE_binary) +
  labs(x = "Number of biological children",
       y = "Estimated slope of HV change (cm3)",
       color = expression(italic("APOE")*"-ε4"),
     fill = expression(italic("APOE")*"-ε4")) +
  th +
  # geom_text(data = data.frame(APOE_binary = factor("Non-carriers", 
  #                                                    levels = c("Non-carriers", "Carriers")), 
  #                             kids_total = Inf, 
  #                             diff_mri.trend = Inf, 
  #                             label = "p=0.257"),
  #           aes(x = kids_total, y = diff_mri.trend, label = label),
  #           size = 4, vjust = 3, hjust = 1.2, inherit.aes = FALSE)+
  # geom_text(data = data.frame(APOE_binary = factor("Carriers", 
  #                                                    levels = c("Non-carriers", "Carriers")), 
  #                             kids_total = Inf, 
  #                             diff_mri.trend = Inf, 
  #                             label = "p=0.001**"),
  #           aes(x = kids_total, y = diff_mri.trend, label = label),
  #           size = 4, vjust = 3, hjust = 1.2, inherit.aes = FALSE)+
  scale_color_manual(values = c("#1CA8B7", "#FF7043"),
                     labels = c("Non-Carriers", "Carriers"))+
  scale_fill_manual(values = c("#1CA8B7", "#FF7043"),
                     labels = c("Non-Carriers", "Carriers"))+
  scale_x_continuous(
    breaks = c(0, 1, 2, 3, 4),
    labels = c("0", "1", "2", "3", "4+")
  )+
  scale_linetype_manual(
    name   = "Model",
    values = c("Unadjusted" = "solid", "Adjusted" = "dashed")
  )+
  guides(
  linetype = guide_legend(
    keywidth = unit(2, "cm")  # increase this until enough dashes show
  )
)

```

```{r, echo=F, warning=F}
p1/p2+
  plot_annotation(tag_levels = 'A')
ggsave("figure1.png", width = 8, height = 7)
```

## Figure 2

```{r, echo=F, warning=F}
# Standardize outcome variables based on visit 1 values
# First, calculate mean and SD from visit 1 only



# Fit models with standardized outcomes - APOE = 0
models_apoe0 <- list(
  CA1 = lmer(vol_CA1_std ~ kids_total2*diff_mri + vol_tiv + age_v1 + education +HRT+ (1 | subject),
             data = filter(data_ashs_female, APOE_binary == 0)),
  CA2 = lmer(vol_CA2_std ~ kids_total2*diff_mri + vol_tiv + age_v1 + education +HRT+ (1 | subject),
             data = filter(data_ashs_female, APOE_binary == 0)),
  CA3 = lmer(vol_CA3_std ~ kids_total2*diff_mri + vol_tiv + age_v1 + education +HRT+ (1 | subject),
             data = filter(data_ashs_female, APOE_binary == 0)),
  DG = lmer(vol_DG_std ~ kids_total2*diff_mri + vol_tiv + age_v1 + education +HRT+ (1 | subject),
            data = filter(data_ashs_female, APOE_binary == 0)),
  SUB = lmer(vol_SUB_std ~ kids_total2*diff_mri + vol_tiv + age_v1 + education +HRT+ (1 | subject),
             data = filter(data_ashs_female, APOE_binary == 0))
)

# Fit models with standardized outcomes - APOE = 1
models_apoe1 <- list(
  CA1 = lmer(vol_CA1_std ~ kids_total2*diff_mri + vol_tiv + age_v1 + education +HRT+ (1 | subject),
             data = filter(data_ashs_female, APOE_binary == 1)),
  CA2 = lmer(vol_CA2_std ~ kids_total2*diff_mri + vol_tiv + age_v1 + education +HRT+ (1 | subject),
             data = filter(data_ashs_female, APOE_binary == 1)),
  CA3 = lmer(vol_CA3_std ~ kids_total2*diff_mri + vol_tiv + age_v1 + education +HRT+ (1 | subject),
             data = filter(data_ashs_female, APOE_binary == 1)),
  DG = lmer(vol_DG_std ~ kids_total2*diff_mri + vol_tiv + age_v1 + education +HRT+ (1 | subject),
            data = filter(data_ashs_female, APOE_binary == 1)),
  SUB = lmer(vol_SUB_std ~ kids_total2*diff_mri + vol_tiv + age_v1 + education +HRT+ (1 | subject),
             data = filter(data_ashs_female, APOE_binary == 1))
)

# Extract coefficients for the interaction term
extract_interaction <- function(model_list, apoe_status) {
  map_df(names(model_list), function(region) {
    model <- model_list[[region]]
    coef_df <- tidy(model, conf.int = TRUE) %>%
      filter(term == "kids_total2:diff_mri") %>%
      mutate(region = region,
             APOE = apoe_status)
    return(coef_df)
  })
}

results_apoe0 <- extract_interaction(models_apoe0, "Non-carriers")
results_apoe1 <- extract_interaction(models_apoe1, "Carriers")

# Combine results
forest_data <- bind_rows(results_apoe0, results_apoe1) %>%
  mutate(
    region = factor(region, levels = rev(c("CA1", "CA2", "CA3", "DG", "SUB"))),
    # Add significance stars
    sig_label = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01 ~ "**",
      p.value < 0.05 ~ "*",
      TRUE ~ ""
    )
  )

# Create forest plot with standardized estimates
forest1<-ggplot(forest_data, aes(x = estimate, y = region, color = APOE)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), 
                 height = 0.2, position = position_dodge(width = 0.5)) +
  geom_point(size = 3, position = position_dodge(width = 0.5)) +
  geom_text(aes(label = sig_label), 
            position = position_dodge(width = 0.5),
            vjust = -0.8, size = 4, show.legend = FALSE) +
  labs(x = "Parity:Time",
       y = "Hippocampal subfields",
       color = expression(italic("APOE")*"-ε4")) +
  th+
  theme(legend.position = "none")
  scale_color_manual(values = c("#1CA8B7", "#FF7043"),
                     labels = c("Non-Carriers", "Carriers"))
```

```{r, echo=F, warning=F}

# Fit models with standardized outcomes - APOE = 0
models_apoe0 <- list(
  CA1 = lmer(vol_CA1_std ~ kids_total2*diff_mri + vol_tiv + age_v1 + education + HRT+CSF_Abeta_ratio+ CSF_Abeta_ratio:diff_mri+ (1 | subject),
             data = filter(data_ashs_female, APOE_binary == 0)),
  CA2 = lmer(vol_CA2_std ~ kids_total2*diff_mri + vol_tiv + age_v1 + education + HRT+CSF_Abeta_ratio+ CSF_Abeta_ratio:diff_mri+ (1 | subject),
             data = filter(data_ashs_female, APOE_binary == 0)),
  CA3 = lmer(vol_CA3_std ~ kids_total2*diff_mri + vol_tiv + age_v1 + education + HRT+CSF_Abeta_ratio+ CSF_Abeta_ratio:diff_mri+ (1 | subject),
             data = filter(data_ashs_female, APOE_binary == 0)),
  DG = lmer(vol_DG_std ~ kids_total2*diff_mri + vol_tiv + age_v1 + education + HRT+CSF_Abeta_ratio+ CSF_Abeta_ratio:diff_mri+ (1 | subject),
            data = filter(data_ashs_female, APOE_binary == 0)),
  SUB = lmer(vol_SUB_std ~ kids_total2*diff_mri + vol_tiv + age_v1 + education + HRT+CSF_Abeta_ratio+ CSF_Abeta_ratio:diff_mri+ (1 | subject),
             data = filter(data_ashs_female, APOE_binary == 0))
)

# Fit models with standardized outcomes - APOE = 1
models_apoe1 <- list(
  CA1 = lmer(vol_CA1_std ~ kids_total2*diff_mri + vol_tiv + age_v1 + education + HRT+CSF_Abeta_ratio+ CSF_Abeta_ratio:diff_mri+ (1 | subject),
             data = filter(data_ashs_female, APOE_binary == 1)),
  CA2 = lmer(vol_CA2_std ~ kids_total2*diff_mri + vol_tiv + age_v1 + education + HRT+CSF_Abeta_ratio+ CSF_Abeta_ratio:diff_mri+ (1 | subject),
             data = filter(data_ashs_female, APOE_binary == 1)),
  CA3 = lmer(vol_CA3_std ~ kids_total2*diff_mri + vol_tiv + age_v1 + education + HRT+CSF_Abeta_ratio+ CSF_Abeta_ratio:diff_mri+ (1 | subject),
             data = filter(data_ashs_female, APOE_binary == 1)),
  DG = lmer(vol_DG_std ~ kids_total2*diff_mri + vol_tiv + age_v1 + education + HRT+CSF_Abeta_ratio+ CSF_Abeta_ratio:diff_mri+ (1 | subject),
            data = filter(data_ashs_female, APOE_binary == 1)),
  SUB = lmer(vol_SUB_std ~ kids_total2*diff_mri + vol_tiv + age_v1 + education + HRT+CSF_Abeta_ratio+ CSF_Abeta_ratio:diff_mri+ (1 | subject),
             data = filter(data_ashs_female, APOE_binary == 1))
)

# Extract coefficients for the interaction term
extract_interaction <- function(model_list, apoe_status) {
  map_df(names(model_list), function(region) {
    model <- model_list[[region]]
    coef_df <- tidy(model, conf.int = TRUE) %>%
      filter(term == "kids_total2:diff_mri") %>%
      mutate(region = region,
             APOE = apoe_status)
    return(coef_df)
  })
}

results_apoe0 <- extract_interaction(models_apoe0, "Non-Carriers")
results_apoe1 <- extract_interaction(models_apoe1, "Carriers")

# Combine results
forest_data <- bind_rows(results_apoe0, results_apoe1) %>%
  mutate(
    region = factor(region, levels = rev(c("CA1", "CA2", "CA3", "DG", "SUB"))),
    # Add significance stars
    sig_label = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01 ~ "**",
      p.value < 0.05 ~ "*",
      TRUE ~ ""
    )
  )

# Create forest plot with standardized estimates
forest2<-ggplot(forest_data, aes(x = estimate, y = region, color = APOE)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), 
                 height = 0.2, position = position_dodge(width = 0.5)) +
  geom_point(size = 3, position = position_dodge(width = 0.5)) +
  geom_text(aes(label = sig_label), 
            position = position_dodge(width = 0.5),
            vjust = -0.8, size = 4, show.legend = FALSE) +
  labs(x = "Parity:Time adjusted",
       y = "Hippocampal subfields",
       color = expression(italic("APOE")*"-ε4")) +
  th+
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.line.y = element_blank() # Removes the axis line itself (depending on the theme used)
  )
  scale_color_manual(values = c("#1CA8B7", "#FF7043"),
                     labels = c("Non-Carriers", "Carriers"))
```

```{r, echo=F, warning=FALSE}
forest1+forest2
ggsave("figure2.png", width = 8, height = 7)
```

# Appendix

## MRI data acquisition

MRI scans were acquired on a in-house 3.0 T scanner (GE Discovery MR750 W 3 T) using the same protocol for all participants, which included one T1- and three T2- weighted sequences. The 3D-T1w sequence was designed with an isotropic voxel size of1mm3 and a matrix size of256x256x160 (TR/TE/TI = 8.0/3.7/450 ms, NSA = 1, flip angle = 8°). Three 3D-T2w sequences, with a voxel size of1 mm× 1 mm× 3 mm, were also used: fluid attenuation inversion recovery (FLAIR: TR/TE/ TI = 11,000/90/2600 ms, flip angle = 160°), fast spin echo (FSE: TR/TE = 5000/85 ms, flip angle = 110°) and, gradient echo (GRE: TR/TE = 1300/23 ms, flip angle = 15°). All scans were visually assessed to verify their quality and to detect incidental findings by a trained neuroradiologist and have been reported elsewhere (Brugulat-Serrat et al. 2017). At visit 1, ten participants were excluded due to the presence of a meningioma, as well as 37 participants due to susceptibility, motion artefacts or segmentation problems, resulting in a total of 561 images available for subsequent analysis. The medial temporal lobe atrophy was assessed by Medial Temporal Atrophy scale [@scheltens1992a].

# References

```{r, include=F}

#not sure you are removing people who don't have more than 1 here
#CSF sTREM2, YKL40, GFAP, and plasma GFAP.

library(fitdistrplus)


gfap<-lmer(CSF_GFAP ~ kids_total*biomarkers_time_diff + age_v1   +(1|subject), data = data_biomarkers_long_female[data_biomarkers_long_female$APOE_binary==1,])

summary(gfap) 

resid_gfap<-residuals(gfap)
fitted_gfap <- fitted(gfap)

bc_gfap <- MASS::boxcox(fitted_gfap + resid_gfap ~ 1)
lambda_gfap <- bc_gfap$x[which.max(bc_gfap$y)]

data_biomarkers_long_female$CSF_GFAP_transformed <- (data_biomarkers_long_female$CSF_GFAP^lambda_gfap - 1) / lambda_gfap

gfap2<-lmer(CSF_GFAP_transformed ~ kids_total*biomarkers_time_diff + age_v1   +(1|subject), data = data_biomarkers_long_female[data_biomarkers_long_female$APOE_binary==1,])
summary(gfap2) 
plot(gfap2)
qqnorm(residuals(gfap2))

##########

strem<-lmer( CSF_sTREM2 ~ kids_total*biomarkers_time_diff + age_v1   +(1|subject), data = data_biomarkers_long_female[data_biomarkers_long_female$APOE_binary==1,])

resid_strem<-residuals(strem)
fitted_strem <- fitted(strem)

bc_strem <- MASS::boxcox(fitted_strem + resid_strem ~ 1)
lambda_strem <- bc_strem$x[which.max(bc_strem$y)]

data_biomarkers_long_female$CSF_sTREM2_transformed <- (data_biomarkers_long_female$CSF_sTREM2^lambda_strem - 1) / lambda_strem

strem2<-lmer(CSF_sTREM2_transformed ~ kids_total*biomarkers_time_diff  +age_v1   +(1|subject), data = data_biomarkers_long_female[data_biomarkers_long_female$APOE_binary==1,])
summary(strem2) 
plot(strem2)
qqnorm(residuals(strem2))

####

yk40<-lmer( CSF_YKL40 ~ kids_total*biomarkers_time_diff + age_v1  + CSF_Abeta_ratio  +(1|subject), data = data_biomarkers_long_female[data_biomarkers_long_female$APOE_binary==1,])

resid_yk40<-residuals(yk40)
fitted_yk40 <- fitted(yk40)

bc_yk40 <- MASS::boxcox(fitted_yk40 + resid_yk40 ~ 1)
lambda_yk40 <- bc_yk40$x[which.max(bc_yk40$y)]

data_biomarkers_long_female$CSF_YKL40_transformed <- (data_biomarkers_long_female$CSF_YKL40^lambda_yk40 - 1) / lambda_yk40

yk402<-lmer(CSF_YKL40_transformed ~ kids_total*biomarkers_time_diff + age_v1 + CSF_Abeta_ratio  +(1|subject), data = data_biomarkers_long_female[data_biomarkers_long_female$APOE_binary==1,])
summary(yk402) 
plot(yk402)
qqnorm(residuals(yk402))


```

```{r, include=F}
svcam1<-lm( CSF_sVCAM1 ~ kids_total + age_v1  + CSF_Abeta_ratio+ education , data = data_biomarkers_long_female[data_biomarkers_long_female$APOE_binary==1 & data_biomarkers_long_female$Visit==1,])

resid_svcam1<-residuals(svcam1)
fitted_svcam1 <- fitted(svcam1)

bc_svcam1 <- MASS::boxcox(fitted_svcam1 + resid_svcam1 ~ 1)
lambda_svcam1 <- bc_svcam1$x[which.max(bc_svcam1$y)]

data_biomarkers_long_female$CSF_sVCAM1_transformed <- (data_biomarkers_long_female$CSF_sVCAM1^lambda_svcam1 - 1) / lambda_svcam1

svcam12<-lm(CSF_sVCAM1_transformed ~ kids_total + CSF_Abeta_ratio+ age_v1  + education , data = data_biomarkers_long_female[data_biomarkers_long_female$APOE_binary==1 & data_biomarkers_long_female$Visit==1,])
summary(svcam12) 
plot(svcam12)
qqnorm(residuals(svcam12))


sicam1<-lm( CSF_sICAM1 ~ kids_total + age_v1  + CSF_Abeta_ratio+ education , data = data_biomarkers_long_female[data_biomarkers_long_female$APOE_binary==1 & data_biomarkers_long_female$Visit==1,])

resid_sicam1<-residuals(sicam1)
fitted_sicam1 <- fitted(sicam1)

bc_sicam1 <- MASS::boxcox(fitted_sicam1 + resid_sicam1 ~ 1)
lambda_sicam1 <- bc_sicam1$x[which.max(bc_sicam1$y)]

data_biomarkers_long_female$CSF_sICAM1_transformed <- (data_biomarkers_long_female$CSF_sICAM1^lambda_sicam1 - 1) / lambda_sicam1

sicam12<-lm(CSF_sICAM1_transformed ~ kids_total + CSF_Abeta_ratio+ age_v1  + education , data = data_biomarkers_long_female[data_biomarkers_long_female$APOE_binary==1 & data_biomarkers_long_female$Visit==1,])
summary(sicam12) 
plot(sicam12)
qqnorm(residuals(sicam12))
```

```{r, include=F, eval=F}
test<-data_ashs_female%>%
   mutate(Visit= ifelse(Visit==1, "V1", "V2"),
          A=as.factor(A))%>%
   left_join(biomarkers)
# 
# mod_ashs_APOE<-lmer(vol_HIPP ~  kids_total*APOE_binary*diff_mri +age_v1 + scale(vol_tiv) + education + CSF_YKL40 +(1|subject), data = test)
# summary(mod_ashs_APOE)

fit_test<-lmer(vol_HIPP ~  CSF_GFAP*diff_mri +age_v1 + scale(vol_tiv) + education +(1|subject), data = test[test$APOE_binary==1,])
summary(fit_test)

resid_fit_test<-residuals(fit_test)
fitted_fit_test <- fitted(fit_test)

bc_fit_test <- MASS::boxcox(fitted_fit_test + resid_fit_test ~ 1)
lambda_fit_test <- bc_fit_test$x[which.max(bc_fit_test$y)]

test$CSF_GFAP_transformed <- (test$CSF_GFAP^lambda_fit_test - 1) / lambda_fit_test

fit_test2<-lmer(vol_HIPP ~  CSF_GFAP_transformed*diff_mri +age_v1 + scale(vol_tiv) + education +(1|subject), data = test[test$APOE_binary==1,])
summary(fit_test2)

##### mediation - time component

test_med <-lmer(vol_HIPP ~  CSF_YKL40_transformed*diff_mri +age_v1 + scale(vol_tiv) + education +(1|subject), data = test[test$APOE_binary==1,])
summary(test_med)


model_mediator <- lmer(CSF_YKL40_transformed ~ kids_total*diff_mri + age_v1 + education +(1|subject), 
                     data = test[test$APOE_binary==1,])

# Model 2: Effect of parity and HV on cognition
model_outcome <- lmer(vol_HIPP ~  kids_total*diff_mri + CSF_YKL40_transformed*diff_mri + scale(vol_tiv) + age_v1 + education +(1|subject), 
                    data = test[test$APOE_binary==1,])

# Check your models
summary(model_mediator)
summary(model_outcome)
set.seed(123)

mediation_results2 <- mediate(model_mediator,    # Mediator model
                              model_outcome,     # Outcome model
                              treat = "kids_total",  # Independent variable
                              mediator = "CSF_YKL40_transformed",   # Mediator
                              boot = FALSE,       # Use bootstrapping
                              sims = 5000) 

summary(mediation_results2)


## mediation - baseline inflammation on HV change
baseline_YKL40<-biomarkers%>%
  filter(Visit=="V1")%>%
  dplyr::select(subject, CSF_YKL40)



vol_tiv_baseline<-data_ashs_female%>%
  filter(Visit==1)%>%
  dplyr::select(subject, vol_tiv)

data_med<-data_ashs_female%>%
  dplyr::select(subject, kids_total, age_v1, diff_mri, vol_HIPP, Visit, APOE_binary, education)%>%
  pivot_wider(names_from = Visit, values_from = c(vol_HIPP, diff_mri))%>%
  mutate(change_HIPP = (vol_HIPP_1+ vol_HIPP_2)/diff_mri_2)%>%
  left_join(baseline_YKL40)%>%
  left_join(., vol_tiv_baseline)

model_mediator_pre <- lm(CSF_YKL40 ~ kids_total + age_v1 + education , 
                     data = data_med[data_med$APOE_binary==1,])
plot(model_mediator_pre)

resid_model_mediator_pre<-residuals(model_mediator_pre)
fitted_model_mediator_pre <- fitted(model_mediator_pre)

bc_model_mediator_pre <- MASS::boxcox(fitted_model_mediator_pre + resid_model_mediator_pre ~ 1)
lambda_model_mediator_pre <- bc_model_mediator_pre$x[which.max(bc_model_mediator_pre$y)]

data_med$CSF_YKL40_transformed <- (data_med$CSF_YKL40^lambda_model_mediator_pre - 1) / lambda_model_mediator_pre

model_mediator<-lm(CSF_YKL40_transformed ~ kids_total + age_v1  + education , data = data_med[data_med$APOE_binary==1,])
plot(model_mediator)

model_outcome <- lm(change_HIPP ~  kids_total + CSF_YKL40_transformed + vol_tiv + age_v1 + education , 
                    data = data_med[data_med$APOE_binary==1,])

mediation_results <- mediate(model_mediator,    # Mediator model
                              model_outcome,     # Outcome model
                              treat = "kids_total",  # Independent variable
                              mediator = "CSF_YKL40_transformed",   # Mediator
                              boot = TRUE,       # Use bootstrapping
                              sims = 5000) 

summary(mediation_results)

### svcam

baseline_svcam<-biomarkers%>%
  filter(Visit=="V1")%>%
  dplyr::select(subject, CSF_sVCAM1)

data_med<-data_med%>%
  left_join(., baseline_svcam)

model_mediator_pre <- lm(CSF_sVCAM1 ~ kids_total + age_v1 + education , 
                     data = data_med[data_med$APOE_binary==1,])
plot(model_mediator_pre)

resid_model_mediator_pre<-residuals(model_mediator_pre)
fitted_model_mediator_pre <- fitted(model_mediator_pre)

bc_model_mediator_pre <- MASS::boxcox(fitted_model_mediator_pre + resid_model_mediator_pre ~ 1)
lambda_model_mediator_pre <- bc_model_mediator_pre$x[which.max(bc_model_mediator_pre$y)]

data_med$CSF_sVCAM1_transformed <- (data_med$CSF_sVCAM1^lambda_model_mediator_pre - 1) / lambda_model_mediator_pre

model_mediator<-lm(CSF_sVCAM1_transformed ~ kids_total + age_v1  + education , data = data_med[data_med$APOE_binary==1,])
plot(model_mediator)

model_outcome <- lm(change_HIPP ~  kids_total + CSF_sVCAM1_transformed + vol_tiv + age_v1 + education , 
                    data = data_med[data_med$APOE_binary==1,])

mediation_results <- mediate(model_mediator,    # Mediator model
                              model_outcome,     # Outcome model
                              treat = "kids_total",  # Independent variable
                              mediator = "CSF_sVCAM1_transformed",   # Mediator
                              boot = TRUE,       # Use bootstrapping
                              sims = 5000) 

summary(mediation_results)

#baseline sicam

baseline_sicam<-biomarkers%>%
  filter(Visit=="V1")%>%
  dplyr::select(subject, CSF_sICAM1)

data_med<-data_med%>%
  left_join(., baseline_sicam)

model_mediator_pre <- lm(CSF_sICAM1 ~ kids_total + age_v1 + education , 
                     data = data_med[data_med$APOE_binary==1,])
plot(model_mediator_pre)

resid_model_mediator_pre<-residuals(model_mediator_pre)
fitted_model_mediator_pre <- fitted(model_mediator_pre)

bc_model_mediator_pre <- MASS::boxcox(fitted_model_mediator_pre + resid_model_mediator_pre ~ 1)
lambda_model_mediator_pre <- bc_model_mediator_pre$x[which.max(bc_model_mediator_pre$y)]

data_med$CSF_sICAM1_transformed <- (data_med$CSF_sICAM1^lambda_model_mediator_pre - 1) / lambda_model_mediator_pre

model_mediator<-lm(CSF_sICAM1_transformed ~ kids_total + age_v1  + education , data = data_med[data_med$APOE_binary==1,])
plot(model_mediator)

model_outcome <- lm(change_HIPP ~  kids_total + CSF_sICAM1_transformed + vol_tiv + age_v1 + education , 
                    data = data_med[data_med$APOE_binary==1,])

mediation_results <- mediate(model_mediator,    # Mediator model
                              model_outcome,     # Outcome model
                              treat = "kids_total",  # Independent variable
                              mediator = "CSF_sICAM1_transformed",   # Mediator
                              boot = TRUE,       # Use bootstrapping
                              sims = 5000) 

summary(mediation_results)

```

```{r, includ=F, eval=F}

library(mice)
# Step 1: Set up and run imputation
pred <- make.predictorMatrix(data_ashs_female)
pred[, "subject"] <- -2
pred["subject", ] <- 0

method_vector <- make.method(data_ashs_female)
method_vector[names(data_ashs_female) != "CSF_Abeta_ratio"] <- ""
method_vector["CSF_Abeta_ratio"] <- "pmm"

imp_data <- mice(data_ashs_female, 
                 m = 5, 
                 method = method_vector, 
                 predictorMatrix = pred,
                 maxit = 50, 
                 seed = 500)

# Step 2: Fit model across all imputations (this is the correct way)
fit <- with(imp_data, 
            lmer(vol_HIPP ~ kids_total*APOE_binary*diff_mri + CSF_Abeta_ratio+
                 age_v1 + scale(vol_tiv) + education + 
                 (1|subject)))

# Step 3: Pool results using Rubin's rules
pooled <- pool(fit)
summary(pooled)

# If you want a tidier output
library(broom.mixed)
tidy(pooled, conf.int = TRUE)
```

```{r, jacks, include=F, eval=F}
jack1<-read_excel(paste0(config$alfa_dataset_neuroimaging, "Jack signature/bx_signature_jack_ALFA_PLUS_V1_20231120_20250527_142751.xlsx"))%>%
  filter(weighted == "TRUE")%>%
  dplyr::select(-ID, -weighted, -signature)%>%
  pivot_wider(values_from = value, names_from = measurement)%>%
  mutate(Visit = 1)

jack2<-read_excel(paste0(config$alfa_dataset_neuroimaging, "Jack signature/bx_signature_jack_ALFA_PLUS_V2_20230518_20250210_114442.xlsx"))%>%
  filter(weighted == "TRUE")%>%
  dplyr::select(-ID, -weighted, -signature)%>%
  pivot_wider(values_from = value, names_from = measurement)%>%
  mutate(Visit = 2)

data_jack<-bind_rows(jack1, jack2)%>%
  mutate(subject=as.numeric(subject))%>%
  right_join(.,data_naked)%>%
  left_join(., diff_mri)

mod_jack<-lmer(GrayVol ~  kids_total*diff_mri*APOE_binary +age_v1 + education  +(1|subject), data = data_jack[data_jack$sex==2, ])
summary(mod_jack)
```

```{r graph subfields, echo=F, warning=F, include=F, eval=F}
preds_CA1 <- ggpredict(mod_ashs_CA1, terms = c("diff_mri", "kids_total", "APOE_binary"))
preds_CA2 <- ggpredict(mod_ashs_CA2, terms = c("diff_mri", "kids_total", "APOE_binary"))
preds_SUB <- ggpredict(mod_ashs_SUB, terms = c("diff_mri", "kids_total", "APOE_binary"))
preds_ERC <- ggpredict(mod_ashs_ERC, terms = c("diff_mri", "kids_total", "APOE_binary"))
preds_DG <- ggpredict(mod_ashs_DG, terms = c("diff_mri", "kids_total", "APOE_binary"))

preds_CA1_A0<-as.data.frame(preds_CA1)%>%
  rename(kids_total_factor = group,
         APOE_binary=facet)%>%
  filter(APOE_binary==0)%>%
  mutate(APOE_binary = "Non-Carriers",,
         kids_total_factor = ifelse(kids_total_factor == 4, "4+", as.character(kids_total_factor)))


preds_CA1_A1<-as.data.frame(preds_CA1)%>%
  rename(kids_total_factor = group,
         APOE_binary=facet)%>%
  filter(APOE_binary==1)%>%
  mutate(APOE_binary = "Carriers",
         kids_total_factor = ifelse(kids_total_factor == 4, "4+", as.character(kids_total_factor)))

preds_CA2_A0<-as.data.frame(preds_CA2)%>%
  rename(kids_total_factor = group,
         APOE_binary=facet)%>%
  filter(APOE_binary==0)%>%
  mutate(APOE_binary = "Non-Carriers",,
         kids_total_factor = ifelse(kids_total_factor == 4, "4+", as.character(kids_total_factor)))


preds_CA2_A1<-as.data.frame(preds_CA2)%>%
  rename(kids_total_factor = group,
         APOE_binary=facet)%>%
  filter(APOE_binary==1)%>%
  mutate(APOE_binary = "Carriers",
         kids_total_factor = ifelse(kids_total_factor == 4, "4+", as.character(kids_total_factor)))

preds_SUB_A0<-as.data.frame(preds_SUB)%>%
  rename(kids_total_factor = group,
         APOE_binary=facet)%>%
  filter(APOE_binary==0)%>%
  mutate(APOE_binary = "Non-Carriers",,
         kids_total_factor = ifelse(kids_total_factor == 4, "4+", as.character(kids_total_factor)))


preds_SUB_A1<-as.data.frame(preds_SUB)%>%
  rename(kids_total_factor = group,
         APOE_binary=facet)%>%
  filter(APOE_binary==1)%>%
  mutate(APOE_binary = "Carriers",
         kids_total_factor = ifelse(kids_total_factor == 4, "4+", as.character(kids_total_factor)))

preds_ERC_A0<-as.data.frame(preds_ERC)%>%
  rename(kids_total_factor = group,
         APOE_binary=facet)%>%
  filter(APOE_binary==0)%>%
  mutate(APOE_binary = "Non-Carriers",,
         kids_total_factor = ifelse(kids_total_factor == 4, "4+", as.character(kids_total_factor)))


preds_ERC_A1<-as.data.frame(preds_ERC)%>%
  rename(kids_total_factor = group,
         APOE_binary=facet)%>%
  filter(APOE_binary==1)%>%
  mutate(APOE_binary = "Carriers",
         kids_total_factor = ifelse(kids_total_factor == 4, "4+", as.character(kids_total_factor)))

preds_DG_A0<-as.data.frame(preds_DG)%>%
  rename(kids_total_factor = group,
         APOE_binary=facet)%>%
  filter(APOE_binary==0)%>%
  mutate(APOE_binary = "Non-Carriers",,
         kids_total_factor = ifelse(kids_total_factor == 4, "4+", as.character(kids_total_factor)))


preds_DG_A1<-as.data.frame(preds_DG)%>%
  rename(kids_total_factor = group,
         APOE_binary=facet)%>%
  filter(APOE_binary==1)%>%
  mutate(APOE_binary = "Carriers",
         kids_total_factor = ifelse(kids_total_factor == 4, "4+", as.character(kids_total_factor)))


CA1_plot<-data_ashs %>%
  filter(sex == 2, !is.na(APOE_binary)) %>%
  mutate(kids_total_factor = ifelse(kids_total == 4, "4+", as.character(kids_total)),
         APOE_binary = factor(APOE_binary, levels = c("0", "1"), labels = c("Non-Carriers", "Carriers"))) %>%
  ggplot(aes(x = diff_mri, y = vol_CA1, color = kids_total_factor)) +
  geom_line(data =preds_CA1_A0, aes(x=x, 
                                y=predicted, 
                                group=kids_total_factor), size=1.3) +
  geom_line(data =preds_CA1_A1, aes(x=x, 
                                y=predicted, 
                                group=kids_total_factor), size=1.3) +
  geom_ribbon(data =preds_CA1_A0, aes(x=x,  
                                  ymin= conf.low, 
                                  ymax = conf.high, 
                                  group=kids_total_factor), inherit.aes = F, alpha=.07)+
  geom_ribbon(data =preds_CA1_A1, aes(x=x,  
                                  ymin= conf.low, 
                                  ymax = conf.high, 
                                  group=kids_total_factor), inherit.aes = F, alpha=.07)+
  facet_wrap(~APOE_binary) +
  scale_color_manual(values = c("0" = "#8AD9B5", "1" = "#50B88E", "2" = "#047A56", "3" = "#0A5940", "4+" = "#0F3F2D")) +
  th +
  labs(x = "Time difference (years)", 
       y = "CA1", 
       color = "Number of childbirths")+
  theme(legend.position = "none")

CA2_plot<-data_ashs %>%
  filter(sex == 2, !is.na(APOE_binary)) %>%
  mutate(kids_total_factor = ifelse(kids_total == 4, "4+", as.character(kids_total)),
         APOE_binary = factor(APOE_binary, levels = c("0", "1"), labels = c("Non-Carriers", "Carriers"))) %>%
  ggplot(aes(x = diff_mri, y = vol_CA2, color = kids_total_factor)) +
  geom_line(data =preds_CA2_A0, aes(x=x, 
                                y=predicted, 
                                group=kids_total_factor), size=1.3) +
  geom_line(data =preds_CA2_A1, aes(x=x, 
                                y=predicted, 
                                group=kids_total_factor), size=1.3) +
  geom_ribbon(data =preds_CA2_A0, aes(x=x,  
                                  ymin= conf.low, 
                                  ymax = conf.high, 
                                  group=kids_total_factor), inherit.aes = F, alpha=.07)+
  geom_ribbon(data =preds_CA2_A1, aes(x=x,  
                                  ymin= conf.low, 
                                  ymax = conf.high, 
                                  group=kids_total_factor), inherit.aes = F, alpha=.07)+
  facet_wrap(~APOE_binary) +
  scale_color_manual(values = c("0" = "#8AD9B5", "1" = "#50B88E", "2" = "#047A56", "3" = "#0A5940", "4+" = "#0F3F2D")) +
  th +
  labs(x = "Time difference (years)", 
       y = "CA2", 
       color = "Number of childbirths")+
  theme(legend.position = "none")

SUB_plot<-data_ashs %>%
  filter(sex == 2, !is.na(APOE_binary)) %>%
  mutate(kids_total_factor = ifelse(kids_total == 4, "4+", as.character(kids_total)),
         APOE_binary = factor(APOE_binary, levels = c("0", "1"), labels = c("Non-Carriers", "Carriers"))) %>%
  ggplot(aes(x = diff_mri, y = vol_SUB, color = kids_total_factor)) +
  geom_line(data =preds_SUB_A0, aes(x=x, 
                                y=predicted, 
                                group=kids_total_factor), size=1.3) +
  geom_line(data =preds_SUB_A1, aes(x=x, 
                                y=predicted, 
                                group=kids_total_factor), size=1.3) +
  geom_ribbon(data =preds_SUB_A0, aes(x=x,  
                                  ymin= conf.low, 
                                  ymax = conf.high, 
                                  group=kids_total_factor), inherit.aes = F, alpha=.07)+
  geom_ribbon(data =preds_SUB_A1, aes(x=x,  
                                  ymin= conf.low, 
                                  ymax = conf.high, 
                                  group=kids_total_factor), inherit.aes = F, alpha=.07)+
  facet_wrap(~APOE_binary) +
  scale_color_manual(values = c("0" = "#8AD9B5", "1" = "#50B88E", "2" = "#047A56", "3" = "#0A5940", "4+" = "#0F3F2D")) +
  th +
  labs(x = "Time difference (years)", 
       y = "SUB", 
       color = "Number of childbirths")+
  theme(legend.position = "none")

ERC_plot<-data_ashs %>%
  filter(sex == 2, !is.na(APOE_binary)) %>%
  mutate(kids_total_factor = ifelse(kids_total == 4, "4+", as.character(kids_total)),
         APOE_binary = factor(APOE_binary, levels = c("0", "1"), labels = c("Non-Carriers", "Carriers"))) %>%
  ggplot(aes(x = diff_mri, y = vol_ERC, color = kids_total_factor)) +
  geom_line(data =preds_ERC_A0, aes(x=x, 
                                y=predicted, 
                                group=kids_total_factor), size=1.3) +
  geom_line(data =preds_ERC_A1, aes(x=x, 
                                y=predicted, 
                                group=kids_total_factor), size=1.3) +
  geom_ribbon(data =preds_ERC_A0, aes(x=x,  
                                  ymin= conf.low, 
                                  ymax = conf.high, 
                                  group=kids_total_factor), inherit.aes = F, alpha=.07)+
  geom_ribbon(data =preds_ERC_A1, aes(x=x,  
                                  ymin= conf.low, 
                                  ymax = conf.high, 
                                  group=kids_total_factor), inherit.aes = F, alpha=.07)+
  facet_wrap(~APOE_binary) +
  scale_color_manual(values = c("0" = "#8AD9B5", "1" = "#50B88E", "2" = "#047A56", "3" = "#0A5940", "4+" = "#0F3F2D")) +
  th +
  labs(x = "Time difference (years)", 
       y = "ERC", 
       color = "Number of childbirths")+
  theme(legend.position = "none")

DG_plot<-data_ashs %>%
  filter(sex == 2, !is.na(APOE_binary)) %>%
  mutate(kids_total_factor = ifelse(kids_total == 4, "4+", as.character(kids_total)),
         APOE_binary = factor(APOE_binary, levels = c("0", "1"), labels = c("Non-Carriers", "Carriers"))) %>%
  ggplot(aes(x = diff_mri, y = vol_DG, color = kids_total_factor)) +
  geom_line(data =preds_DG_A0, aes(x=x, 
                                y=predicted, 
                                group=kids_total_factor), size=1.3) +
  geom_line(data =preds_DG_A1, aes(x=x, 
                                y=predicted, 
                                group=kids_total_factor), size=1.3) +
  geom_ribbon(data =preds_DG_A0, aes(x=x,  
                                  ymin= conf.low, 
                                  ymax = conf.high, 
                                  group=kids_total_factor), inherit.aes = F, alpha=.07)+
  geom_ribbon(data =preds_DG_A1, aes(x=x,  
                                  ymin= conf.low, 
                                  ymax = conf.high, 
                                  group=kids_total_factor), inherit.aes = F, alpha=.07)+
  facet_wrap(~APOE_binary) +
  scale_color_manual(values = c("0" = "#8AD9B5", "1" = "#50B88E", "2" = "#047A56", "3" = "#0A5940", "4+" = "#0F3F2D")) +
  th +
  labs(x = "Time difference (years)", 
       y = "DG", 
       color = "Number of childbirths")
  


(CA1_plot + CA2_plot ) / 
(ERC_plot + SUB_plot) 
```

```{r graph whole hippocampus, eval=F, include=F}

preds_ashs <- ggpredict(mod_ashs_APOE, terms = c("diff_mri", "kids_total", "APOE_binary"))

preds_ashs_A0<-as.data.frame(preds_ashs)%>%
  rename(kids_total_factor = group,
         APOE_binary=facet)%>%
  filter(APOE_binary==0)%>%
  mutate(APOE_binary = "Non-Carrier",
         kids_total_factor= ifelse(as.character(kids_total_factor)==4, "4+", as.character(kids_total_factor)))

preds_ashs_A1<-as.data.frame(preds_ashs)%>%
  rename(kids_total_factor = group,
         APOE_binary=facet)%>%
  filter(APOE_binary==1)%>%
  mutate(APOE_binary = "Carrier",
         kids_total_factor= ifelse(as.character(kids_total_factor)==4, "4+", as.character(kids_total_factor)))

# Create a combined factor for color mapping
data_plot <- data_ashs %>%
  filter(sex == 2, !is.na(APOE_binary)) %>%
  mutate(kids_total_factor = ifelse(kids_total==4, "4+", kids_total),
         APOE_binary = factor(APOE_binary, levels = c("0", "1"), labels = c("Non-Carrier", "Carrier")),
         color_group = paste(APOE_binary, kids_total_factor, sep = "_"))

preds_ashs_A0 <- preds_ashs_A0 %>%
  mutate(color_group = paste("Non-Carrier", kids_total_factor, sep = "_"))

preds_ashs_A1 <- preds_ashs_A1 %>%
  mutate(color_group = paste("Carrier", kids_total_factor, sep = "_"))


# Define color palettes
color_values <- c(
  "Non-Carrier_0" = "#4DD0E1", "Non-Carrier_1" = "#1CA8B7", "Non-Carrier_2" = "#1E8895", "Non-Carrier_3" = "#0D707B", "Non-Carrier_4+" = "#0F565E",
  "Carrier_0" = "#FF8A65", "Carrier_1" = "#FF7043", "Carrier_2" = "#FF5722", "Carrier_3" = "#F4511E", "Carrier_4+" = "#BF360C"
)

ggplot(data_plot, aes(x = diff_mri, y = vol_HIPP, color = color_group)) +
  geom_point(alpha = 0.3, size=2) +
  geom_line(data = preds_ashs_A0, aes(x=x, y=predicted, group=kids_total_factor, color=color_group), size=1.3) +
  geom_line(data = preds_ashs_A1, aes(x=x, y=predicted, group=kids_total_factor, color=color_group), size=1.3) +
  geom_ribbon(data = preds_ashs_A0, aes(x=x, ymin=conf.low, ymax=conf.high, 
                                         group=kids_total_factor, fill=color_group), 
              inherit.aes = F, alpha=.15)+
  geom_ribbon(data = preds_ashs_A1, aes(x=x, ymin=conf.low, ymax=conf.high, 
                                         group=kids_total_factor, fill=color_group), 
              inherit.aes = F, alpha=.15)+
  facet_wrap(~APOE_binary) +
  scale_color_manual(values = color_values, labels = c("0", "1", "2", "3", "4+", "0", "1", "2", "3", "4+")) +
  scale_fill_manual(values = color_values, labels = c("0", "1", "2", "3", "4+", "0", "1", "2", "3", "4+")) +
  labs(x = "Time difference (days)", 
       y = "PACC Z-score", 
       color = "Number of childbirths",
       fill = "Number of childbirths")+
  th
```

```{r power analysis, include=F, eval=F}

#power analysis if needed
sd_kids_total <- sd(data_ashs$kids_total[data_ashs$sex==2], na.rm=TRUE)
prop_A <- prop.table(table(data_ashs$APOE_binary[data_ashs$sex==2]))
p_A <- 0.488 
sd_A <- sqrt(p_A * (1 - p_A))

(34.677 * (sd_kids_total * sd_A) / sd(data_ashs$vol_HIPP[data_ashs$sex==2], na.rm=TRUE))^2

power.t.regression(beta = 34.677,
                   sd.predictor = sd_kids_total * sd_A,
                   sd.outcome = sd(data_ashs$vol_HIPP[data_ashs$sex==2], na.rm=TRUE),
                   r.squared = (34.677 * (sd_kids_total * sd_A) / sd(data_ashs$vol_HIPP[data_ashs$sex==2], na.rm=TRUE))^2,
                   null.beta = 0, 
                   k.total = 8, 
                   n=196, 
                   alpha=0.05)

```

```{r test , eval = F, include = F}
vif(mod_ashs_APOE)

mod_ashs_APOE<-lmer(vol_HIPP ~  kids_total*APOE_binary*diff_mri +age_v1 + scale(vol_tiv) + education  +(1|subject), data = data_ashs[data_ashs$sex==2 & data_ashs$A==0,])
summary(mod_ashs_APOE)


plot_model(mod_ashs_APOE, type = "int")


data_ashs_wide <- data_ashs %>%
  dplyr::select(subject, sex, age_v1, kids_total, APOE_binary, education, starts_with("vol_"), Visit, A, CSF_Abeta_ratio) %>%
  drop_na(Visit) %>%
  pivot_wider(names_from = Visit, values_from = starts_with("vol_")) %>%
  left_join(., date_mri_long)



data_ashs_wide <- data_ashs %>%
  dplyr::select(subject, sex, age_v1, kids_total, APOE_binary, education, starts_with("vol_"), Visit, A, CSF_Abeta_ratio) %>%
  drop_na(Visit) %>%
  pivot_wider(names_from = Visit, values_from = starts_with("vol_")) %>%
  left_join(., date_mri_long) %>%
  {
    df <- .
    vol_1_cols <- grep("^vol_.*_1$", names(df), value = TRUE)
    
    for (col_1 in vol_1_cols) {
      base_name <- sub("_1$", "", col_1)
      col_2 <- paste0(base_name, "_2")
      
      if (col_2 %in% names(df)) {
        df[[paste0(base_name, "_diff")]] <- (df[[col_2]] - df[[col_1]]) / df$diff_mri
      }
    }
    df
  }


test<-lm(vol_HIPP_diff ~ kids_total*APOE_binary + age_v1 + education + scale(vol_tiv_2), data = data_ashs_wide[data_ashs_wide$sex==2 & data_ashs_wide$A==0,])

summary(test)

data_ashs%>%
  filter(APOE_binary==1, A==1, sex==2)%>%
  dplyr::select(subject)%>%
  unique%>%
  nrow()

data_ashs <- data_ashs %>%
  mutate(apoeAB = case_when(
    APOE_binary == 0 & A == 0 ~ 1,
    APOE_binary == 0 & A == 1 ~ 2,
    APOE_binary == 1 & A == 0 ~ 3,
    APOE_binary == 1 & A == 1 ~ 4
  ))%>%
  mutate(apoeAB = factor(apoeAB,levels = c(1,2,3,4)))

test1<-lmer(vol_HIPP ~ kids_total*apoeAB*diff_mri + age_v1 + education + scale(vol_tiv) + (1|subject), data = data_ashs[data_ashs$sex==2,])

# Get predicted values
pred <- ggpredict(test1, terms = c("kids_total [all]", "apoeAB [1,2,3,4]", "diff_mri"))

# Plot with custom colors
ggplot(pred, aes(x = x, y = predicted, color = factor(group), linetype = factor(facet))) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = factor(group)), 
              alpha = 0.1, color = NA) +
  scale_color_manual(values = c("#1b9e77", "#d95f02", "#7570b3", "#e7298a"),
                     labels = c("APOE4-/AB-", "APOE4-/AB+", "APOE4+/AB-", "APOE4+/AB+")) +
  scale_fill_manual(values = c("#1b9e77", "#d95f02", "#7570b3", "#e7298a"),
                    labels = c("APOE4-/AB-", "APOE4-/AB+", "APOE4+/AB-", "APOE4+/AB+")) +
  facet_wrap(~facet, labeller = labeller(facet = c("Low parity", "High parity"))) +
  labs(x = "Time", y = "Hippocampal Volume", 
       color = "Group", fill = "Group") +
  theme_minimal()
```
