---
title: "reproduction APOE4*parity analysis"
author: "Clazz"
date: "2025-11-13"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = F}

if (!require("pacman")) install.packages("pacman")

pacman::p_load(tidyverse, VIM, ggpubr, ggeffects, car, sjPlot,caret, ggrain, emmeans, skimr, lavaan, patchwork, lme4, lmerTest, gtsummary, huxtable)

config <- config::get()

source("C:/Users/U253662/Documents/git/functions/jt_plot_aesthetics.R")
```



# WRAP

not info on gestational diabetes/preeclampsia
CU exclusion is based on a "consensus"
poor availability of amyloid dat hence not included

```{r, include=F}
ashs_WRAP<-read_delim(paste0(config$wrap_dataset_cleaned, "ashs_dataset.csv"),col_select = -1)

menopause<- read_delim(paste0(config$wrap_dataset_cleaned, "menopause_dataset.csv"),col_select = -1)

ptau217<-read_delim(paste0(config$wrap_dataset_cleaned, "ptau217_dataset.csv"),col_select = -1)%>%
  dplyr::select(-days_since_baseline)

consensus<-read_delim("G:/My Drive/Data/WRAP/ConsensusConference.csv")%>%
  mutate(exclude_CU = ifelse(CC2==0, 1, 0))%>%
  rename(subject = WRAPNo)%>%
  dplyr::select(subject, exclude_CU, VisNo)%>%
  rename(visit = VisNo)

WRAP<-read_delim(paste0(config$wrap_dataset_cleaned, "basic_dataset.csv"), col_select = -1)%>%
  left_join(.,ashs_WRAP)%>%
  group_by(subject)%>%
  mutate(age_MR_baseline = first(age_at_acquisition))%>%
  drop_na(kids_total, whole_hippocampus)%>%
  mutate(APOE_binary = factor(APOE_binary, levels = c(0, 1)),
         parous_group = ifelse(kids_total>1, "2+", as.character(kids_total)),
         factor_kids_total = factor(kids_total),
         parous_group2 = ifelse(kids_total>2, "3+", as.character(kids_total)),
         SeriesDescription = as.factor(SeriesDescription))%>%
  filter(sex==2)%>%
  left_join(., menopause)%>%
  mutate(is_menopause = ifelse(age_at_acquisition<age_menopause, 0, 1))%>%
  filter(is_menopause!=0)%>%
  mutate(AcquisitionPlane = case_when(
    str_detect(SeriesDescription, "Ax|axial") ~ "Axial",
    str_detect(SeriesDescription, "Sag|SAG") ~ "Sagittal",
    .default = "Unknown"
  ))%>%
  group_by(subject)%>%
  arrange(age_at_acquisition, .by_group = TRUE) %>% 
  mutate(days_since_first_scan = days_since_baseline - first(days_since_baseline),
         whole_hippocampus = whole_hippocampus/1000,
         visit = row_number(),
         years_since_first_scan = days_since_first_scan/365.25)%>%
  filter(n()>1)%>%
  left_join(., consensus)%>%
  mutate(kids_total2 = ifelse(kids_total == 4, 3, as.numeric(kids_total)),
         ashs_tiv_scaled = scale(ashs_tiv))


WRAP_CU<-WRAP%>%
  filter(exclude_CU==0)%>%
  group_by(subject)%>%
  filter(n()>1)%>%
  ungroup()

# participants<-WRAP%>%
#   filter(sex==2)%>%
#   dplyr::select(subject)%>%
#   unique()%>%
#   pull()
# writeLines(participants, con = "WRAP_participants.txt")
```

## Descriptives WRAP

```{r, echo=F, warning=F}

WRAP_descriptive<-WRAP_CU%>%
  group_by(subject)%>%
  mutate(visit_total = last(visit),
         years_total = last(years_since_first_scan))%>%
  slice(1)%>%
  ungroup()%>%
  dplyr::select(subject, kids_total, APOE_binary, education, age_MR_baseline, whole_hippocampus, visit_total, years_total)%>%
  mutate(APOE_binary=ifelse(APOE_binary==1, "Carrier", "Non-Carrier"),
         kids_total=as.numeric(kids_total))%>%
  dplyr::select(-subject)

desc_table<-tbl_summary(WRAP_descriptive,
                        type = list(visit_total ~ "continuous"),
                        statistic = list(all_continuous() ~ "{mean} ({sd})"),
            by = kids_total, 
            missing = "no",
            label= list(kids_total = "Number of biologcal children",
                     APOE_binary = "APOE-E4 carrier, Count (%)",
                     age_MR_baseline = "Age at first scan, Mean (SD)",
                     education = "Education (years), Mean (SD)",
                     visit_total = "Number of visits",
                     years_total = "Average span between first and last",
                     whole_hippocampus = "Average hippocampus volume at V1, Mean (SD)"
                     ))%>%
  add_n()%>%
  add_p()%>%
  modify_header(label = "**Variables**") %>% # update the column header
  bold_labels()

desc_table

```

-   APOE4 carriership split fairly equal to ALFA!
-   Population profile is not too dismimilar, age at baseline is younger but scans were done later than baseline in most cases (to do: compute age at first scan variable)
-   Hippocampal segmentation is using SPM toolbox, not ASHS, so ROI could be different and quality, and so subfields available

## QC

```{r, echo=F, warning=F, eval=F}
hipp_v1<-WRAP_CU%>%
  group_by(subject)%>%
  filter(n()>1)%>%
  ungroup()%>%
  filter(visit==1)%>%
  pull(whole_hippocampus)

hipp_v2<-WRAP_CU%>%
  filter(visit==2)%>%
  pull(whole_hippocampus)

plot(hipp_v1, hipp_v2)

hipp_v22<-WRAP_CU%>%
  group_by(subject)%>%
  filter(n()>2)%>%
  ungroup()%>%
  filter(visit==2)%>%
  pull(whole_hippocampus)

hipp_v3<-WRAP_CU%>%
  filter(visit==3)%>%
  pull(whole_hippocampus)

plot(hipp_v22, hipp_v3)

hipp_v32<-WRAP_CU%>%
  group_by(subject)%>%
  filter(n()>3)%>%
  ungroup()%>%
  filter(visit==3)%>%
  pull(whole_hippocampus)

hipp_v4<-WRAP_CU%>%
  filter(visit==4)%>%
  pull(whole_hippocampus)

plot(hipp_v32, hipp_v4)

WRAP_CU <- WRAP_CU %>%
  mutate(
    z_score = scale(whole_hippocampus),
    outlier_flag = abs(z_score) > 3
  ) %>%
  ungroup()

WRAP_test <- WRAP_CU %>%
  arrange(subject, visit) %>%
  group_by(subject) %>%
  mutate(
    pct_change = (whole_hippocampus - lag(whole_hippocampus)) / lag(whole_hippocampus) * 100,
    change_flag = abs(pct_change) > 15
  ) %>%
  ungroup()

#participants flagged: 0202, 0424, 0922



```

## Reproducing analysis: APOE x parity (continuous) x time - 2 visits

### Main group analysis 
```{r, echo=F, warning=F}
fit_wrap_2 <-lmer(whole_hippocampus ~ years_since_first_scan*kids_total*APOE_binary + ashs_tiv_scaled + DeviceSerialNumber+ age_MR_baseline + Ethnicity  + (1|subject) , data=WRAP_CU[WRAP_CU$visit<3,])

preds_2 <- ggpredict(fit_wrap_2, terms = c("years_since_first_scan", "kids_total", "APOE_binary"))

huxreg(fit_wrap_2, bold_signif = 0.05, outer_borders =1, stars = c("†" = 0.1, "*" = 0.05, "**" = 0.01, "***" = 0.001), coefs = c("Intercept" = "(Intercept)",
                 "Parity" = "kids_total",
                 "APOE_E4 Carrier" = "APOE_binary1",
                 "Age baseline" = "age_MR_baseline",
                 "TIV" = "ashs_tiv_scaled",
                 "Parity * APOE4" = "kids_total:APOE_binary1",
                 "Parity * time" = "years_since_first_scan:kids_total",
                 "APOE4 * time" = "years_since_first_scan:APOE_binary1",
                 "Parity * APOE4 * time"="years_since_first_scan:kids_total:APOE_binary1"))
```

### APOE neg only
```{r, echo=F, warning=F}
fit_wrap_2_A0 <-lmer(whole_hippocampus ~ years_since_first_scan*kids_total + ashs_tiv_scaled  + age_MR_baseline + DeviceSerialNumber  + Ethnicity + (1|subject), data=WRAP_CU[WRAP_CU$APOE_binary==0 & WRAP_CU$visit<3, ])


huxreg(fit_wrap_2_A0, bold_signif = 0.05, outer_borders =1, stars = c("†" = 0.1, "*" = 0.05, "**" = 0.01, "***" = 0.001), coefs = c("Intercept" = "(Intercept)",
                 "Parity" = "kids_total",
                 "Age baseline" = "age_MR_baseline",
                 "TIV" = "ashs_tiv_scaled",
                 "Parity * time" = "years_since_first_scan:kids_total"))
```

### APOE pos only
```{r, echo=F, warning=F}
fit_wrap_2_A1 <-lmer(whole_hippocampus ~ years_since_first_scan*kids_total + ashs_tiv_scaled  + age_MR_baseline   +DeviceSerialNumber  + Ethnicity+ (1|subject), data=WRAP_CU[WRAP_CU$APOE_binary==1 & WRAP_CU$visit<3, ])

huxreg(fit_wrap_2_A1, bold_signif = 0.05, outer_borders =1, stars = c("†" = 0.1, "*" = 0.05, "**" = 0.01, "***" = 0.001), coefs = c("Intercept" = "(Intercept)",
                 "Parity" = "kids_total",
                 "Age baseline" = "age_MR_baseline",
                 "TIV" = "ashs_tiv_scaled",
                 "Parity * time" = "years_since_first_scan:kids_total"))
```

```{r, echo= F, warning=FALSE}
WRAP_CU_2<- WRAP%>%
  filter(visit <3)

individual_slopes <- WRAP_CU_2 %>%
  filter(!is.na(kids_total) & !is.na(APOE_binary) & !is.na(years_since_first_scan) & !is.na(whole_hippocampus)) %>%
  group_by(subject, kids_total, APOE_binary) %>%  # Only keep subjects with at least 2 observations
  summarise(
    individual_slope = coef(lm(whole_hippocampus ~ years_since_first_scan))[2],
    n_obs = n(),
    .groups = 'drop'
  ) %>%
  filter(!is.na(individual_slope))%>%
  mutate(APOE_binary = as.factor(APOE_binary))



time_range <- range(WRAP_CU_2$years_since_first_scan, na.rm = TRUE)
time_seq <- seq(from = time_range[1], to = time_range[2])

emm_grid <- emmeans(fit_wrap_2, ~ years_since_first_scan * kids_total * APOE_binary,
                    at = list(time = time_seq),
                    data= WRAP_CU_2)
emm_df <- as.data.frame(emm_grid)

slopes <- emtrends(fit_wrap_2, ~ kids_total * APOE_binary, var = "years_since_first_scan", data= WRAP_CU_2, at = list(kids_total = c(0, 1, 2, 3, 4)))
slopes_df <- as.data.frame(slopes)

p1<-ggplot(slopes_df, aes(x = kids_total, y = years_since_first_scan.trend, 
                            fill = as.factor(APOE_binary))) +
  geom_col(position = "dodge", alpha = 0.8, color = "black", size = 0.3) +
  geom_errorbar(aes(ymin = years_since_first_scan.trend - SE, ymax = years_since_first_scan.trend + SE),
                position = position_dodge(width = 0.9), width = 0.3)+
  th+
  labs(x = "Number of childbirths", 
     y = "Time slope", 
     color = expression(italic("APOE")*"-ε4"),
     fill = expression(italic("APOE")*"-ε4"))+
 geom_point(data = individual_slopes, 
             aes(x = kids_total, y = individual_slope, 
                 color = as.factor(APOE_binary)),
             position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.9),
             alpha = 0.4, size = 2, shape=21, colour="black") +
  scale_fill_manual(values = c("#1CA8B7", "#FF7043"),
                     labels = c("Non-Carriers", "Carriers"))+
scale_color_manual(values = c("#1CA8B7", "#FF7043"),
                     labels = c("Non-Carriers", "Carriers"))+
  scale_x_continuous(
    breaks = c(0, 1, 2, 3, 4),
    labels = c("0", "1", "2", "3", "4+")
  )

p1
```


## Reproducing analysis: APOE x parity (continuous) x time - all visits

### Main group analysis 
```{r, echo=F, warning=F}
fit_wrap <-lmer(whole_hippocampus ~ years_since_first_scan*kids_total*APOE_binary + ashs_tiv_scaled + DeviceSerialNumber+ age_MR_baseline + Ethnicity  + (1|subject) , data=WRAP_CU)

preds <- ggpredict(fit_wrap, terms = c("years_since_first_scan", "kids_total", "APOE_binary"))

huxreg(fit_wrap, bold_signif = 0.05, outer_borders =1, stars = c("†" = 0.1, "*" = 0.05, "**" = 0.01, "***" = 0.001), coefs = c("Intercept" = "(Intercept)",
                 "Parity" = "kids_total",
                 "APOE_E4 Carrier" = "APOE_binary1",
                 "Age baseline" = "age_MR_baseline",
                 "TIV" = "ashs_tiv_scaled",
                 "Parity * APOE4" = "kids_total:APOE_binary1",
                 "Parity * time" = "years_since_first_scan:kids_total",
                 "APOE4 * time" = "years_since_first_scan:APOE_binary1",
                 "Parity * APOE4 * time"="years_since_first_scan:kids_total:APOE_binary1"))
```

### APOE neg only
```{r, echo=F, warning=F}
fit_wrap_A0 <-lmer(whole_hippocampus ~ years_since_first_scan*kids_total + ashs_tiv_scaled  + age_MR_baseline + DeviceSerialNumber  + Ethnicity + (1|subject), data=WRAP[WRAP$APOE_binary==0, ])


huxreg(fit_wrap_A0, bold_signif = 0.05, outer_borders =1, stars = c("†" = 0.1, "*" = 0.05, "**" = 0.01, "***" = 0.001), coefs = c("Intercept" = "(Intercept)",
                 "Parity" = "kids_total",
                 "Age baseline" = "age_MR_baseline",
                 "TIV" = "ashs_tiv_scaled",
                 "Parity * time" = "years_since_first_scan:kids_total"))
```

### APOE pos only
```{r, echo=F, warning=F}
fit_wrap_A1 <-lmer(whole_hippocampus ~ years_since_first_scan*kids_total + ashs_tiv_scaled  + age_MR_baseline   +DeviceSerialNumber  + Ethnicity+ (1|subject), data=WRAP[WRAP$APOE_binary==1, ])
huxreg(fit_wrap_A1, bold_signif = 0.05, outer_borders =1, stars = c("†" = 0.1, "*" = 0.05, "**" = 0.01, "***" = 0.001), coefs = c("Intercept" = "(Intercept)",
                 "Parity" = "kids_total",
                 "Age baseline" = "age_MR_baseline",
                 "TIV" = "ashs_tiv_scaled",
                 "Parity * time" = "years_since_first_scan:kids_total"))
```


```{r, echo=F, warning=F}

preds_apoe<-ggemmeans(fit_wrap, 
  terms = c("years_since_first_scan", "kids_total", "APOE_binary"))

preds_apoe0<-as.data.frame(preds_apoe)%>%
  rename(kids_total = group,
         APOE_binary=facet)%>%
  filter(APOE_binary==0)


preds_apoe1<-as.data.frame(preds_apoe)%>%
  rename(kids_total = group,
         APOE_binary=facet)%>%
  filter(APOE_binary==1)
  

ggplot(WRAP_CU, aes(x= years_since_first_scan, y= whole_hippocampus, group= subject, colour = kids_total))+
  geom_line(alpha=.70)+
  geom_line(data = preds_apoe0, aes(x=x, y= predicted, group=kids_total), inherit.aes = F, size=1)+
  geom_ribbon(data = preds_apoe0, aes(x = x, ymin= conf.low, ymax = conf.high, group = kids_total), inherit.aes = F, alpha=0.1)+
  geom_line(data = preds_apoe1, aes(x=x, y= predicted, group=kids_total), inherit.aes = F, size=1)+
  geom_ribbon(data = preds_apoe1, aes(x = x, ymin= conf.low, ymax = conf.high, group = kids_total), inherit.aes = F, alpha=0.1)+
  facet_wrap(~ APOE_binary)+
  th
```






```{r, include=F}


effects_wrap <- ggeffect(fit_wrap, terms = c("years_since_first_scan", "kids_total", "APOE_binary"))

# Create grid of values for prediction - specify multiple time points
time_range <- range(WRAP$years_since_first_scan, na.rm = TRUE)
time_seq <- seq(from = time_range[1], to = time_range[2])

emm_grid <- emmeans(fit_wrap, ~ years_since_first_scan * kids_total * APOE_binary,
                    at = list(time = time_seq),
                    data= WRAP)
emm_df <- as.data.frame(emm_grid)

slopes <- emtrends(fit_wrap, ~ kids_total * APOE_binary, var = "years_since_first_scan", data= WRAP, at = list(kids_total = c(0, 1, 2, 3, 4)))
slopes_df <- as.data.frame(slopes)



individual_slopes <- WRAP %>%
  drop_na(APOE_binary) %>%
  group_by(subject, kids_total, APOE_binary) %>%
  filter(n() >= 2) %>%
  do(model = lm(whole_hippocampus ~ years_since_first_scan, data = .)) %>%
  mutate(slope = coef(model)[2]) %>%
  dplyr::select(-model) %>%
  ungroup()

ggplot(slopes_df, aes(x = kids_total, y = years_since_first_scan.trend, 
                            fill = as.factor(APOE_binary))) +
  geom_col(position = "dodge", alpha = 0.8, color = "black", size = 0.3) +
  geom_errorbar(aes(ymin = years_since_first_scan.trend - SE, ymax = years_since_first_scan.trend + SE),
                position = position_dodge(width = 0.9), width = 0.3)+
  labs(x = "Number of childbirths", 
       y = "Time slope", 
       color = "APOE4 Carriership",
       fill = "APOE4 Carriership",
       title = "Predicted slopes of hippocampal volume by APOE4 carriership and number of kids")+
  scale_fill_manual(values = c("#1CA8B7", "#FF7043"),
                     labels = c("Non Carrier", "Carrier"))+
   geom_point(data = individual_slopes, 
             aes(x = kids_total, y = slope, fill = APOE_binary),
             position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.9),
             alpha = 0.4, size = 1.9, shape = 21)
```





# HABS HD

-   HABS HD neuroimaging data was technically pulled form their website for quality checks 
- ASHS values oddly small and jordi doesn't understand eitehr the slices thing
-   Some incoherences like hughe amount of NAs for TIV compared with available neuroimaging data and 2 different values that cannot be used together or only used as an average which seems odd 
-   Results were null once adjusting for all variables as stated in imaging: scanner, TIV, number of slices 
-   lots of missings once including all covariates 
-   maybe adjusting for social variables would help?

```{r reproduction with HABS, eval=F, include=F}

data_HABS<-read_delim("G:/My Drive/Data/HABS/data_cleaned/basic_dataset.csv", col_select = -1)%>%
  drop_na(kids_total, APOE4_Positivity)%>%
  mutate(parous_group= ifelse(kids_total>1, "2+", kids_total),
         APOE4_Positivity= factor(APOE4_Positivity))
imaging_HABS<-read_delim("G:/My Drive/Data/HABS/imaging/imaging_dataset_long.csv", col_select = -1)%>%
  rename(MRI_date = `01_MRI_Date`,
         centiloids = `01_AB_FBB_Global_SUVR`)
biomarkers_HABS<- read_delim("G:/My Drive/Data/HABS/data_cleaned/biomarkers_long.csv", col_select = -1)%>%
  dplyr::select(subject, Visit_ID, r7_QTX_Plasma_pTau217_PLUS, r7_QTX_Plasma_pTau217_V2)

date_imaging_HABS<-imaging_HABS%>%
  dplyr::select(subject, Visit_ID, MRI_date)%>%
  group_by(subject) %>%  # replace with your actual ID variable
  arrange(subject, MRI_date) %>%  # replace visit_date with your date column
  mutate(
    visit1_date = first(MRI_date),
    time_diff = as.numeric(MRI_date - visit1_date)
  ) %>%
  ungroup()%>%
  dplyr::select(subject, Visit_ID, time_diff)

data_HABS_analysis<-left_join(data_HABS, imaging_HABS, join_by(subject, sex))%>%
  left_join(., date_imaging_HABS)%>%
  mutate(time_diff_scaled = scale(time_diff),
         kids_total=as.numeric(kids_total))%>%
  left_join(., biomarkers_HABS)

data_HABS_analysis_complete<-data_HABS_analysis%>%
  drop_na(vol_HIPP)%>%
  mutate(kids_total_cut = ifelse(kids_total>3, 4, kids_total))%>%
  filter(Visit_ID<3)


habs_fit<-lmer(vol_HIPP ~  kids_total_cut*APOE4_Positivity*time_diff +age_v1 + scanner+ slices_total+ scale(TIV)  + ID_Education + (1|subject), data = data_HABS_analysis_complete)
summary(habs_fit)
plot_model(habs_fit, type = "int")

vif(habs_fit)

library(ggeffects)
effects_habs <- ggeffect(habs_fit, terms = c("time_diff", "kids_total", "APOE4_Positivity"))

# Create grid of values for prediction - specify multiple time points
time_range <- range(data_HABS_analysis$time_diff, na.rm = TRUE)
time_seq <- seq(from = time_range[1], to = time_range[2], length.out = 10)

emm_grid <- emmeans(habs_fit, ~ time_diff * kids_total * APOE4_Positivity,
                    at = list(time = time_seq),
                    data= data_HABS_analysis)
emm_df <- as.data.frame(emm_grid)

slopes <- emtrends(habs_fit, ~ kids_total * APOE4_Positivity, var = "time_diff", data= data_HABS_analysis, at = list(kids_total = c(0, 1, 2, 3, 4, 5, 6)))
slopes_df <- as.data.frame(slopes)


ggplot(slopes_df, aes(x = kids_total, y = time_diff.trend, 
                            fill = as.factor(APOE4_Positivity))) +
  geom_col(position = "dodge", alpha = 0.8, color = "black", size = 0.3) +
  geom_errorbar(aes(ymin = time_diff.trend - SE, ymax = time_diff.trend + SE),
                position = position_dodge(width = 0.9), width = 0.3)+
  th

```
